### 2018-10-27 paper/paper.tex
paper: Remove introduction

\documentclass[10pt,sigconf,authorversion]{lpc}
\usepackage{balance}
\usepackage{courier}
\usepackage{helvet}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{listings}
\usepackage{times}
\usepackage{xcolor}

\pdfinfo{
/Title (Building socket-aware BPF programs)
/Author (Joe Stringer)}

\title{Building socket-aware BPF programs}
\author{Joe Stringer}
\affiliation{%
      \institution{Cilium.io}}
\email{joe@cilium.io}

\acmDOI{}
\setcounter{secnumdepth}{0}
\setcopyright{none}

\usepackage{xcolor}
\newcommand\todo[1]{\textcolor{red}{#1}}

\begin{document}

\begin{abstract}

    Over the past several years, BPF has steadily become more powerful in
    multiple ways: Through building more intelligence into the verifier which
    allows more complex programs to be loaded, and through extension of the API
    such as by adding new map types and new native BPF function calls. While
    BPF has its roots in applying filters at the socket layer, the ability to
    introspect the sockets relating to traffic being filtered has been limited.

    This paper presents the changes necessary to extend socket introspection
    capabilities of the Linux kernel's BPF packet hooks. To provide safety
    guarantees around the access of the socket, the verifier is extended to
    track allocation and release of kernel resources. Using this mechanism, we
    introduce socket lookup and release helpers into the BPF API.

    %These new primitives open up new possibilities for various use cases
    %including stateful firewalling and load balancing.

    %To provide access to socket state from a BPF helper call, the verifier
    %needs the ability to track the safety of the calls, including appropriate
    %reference counting upon the underlying socket for the duration of the
    %access.

    %With such an ability ability in the verifier, we add socket lookup and
    %release function calls to the BPF API, allowing BPF programs to safely find
    %a socket and build logic upon the presence or attributes of a socket.
    %Applications for this functionality range from load-balancing traffic based
    %on the presence of a listening application to building stateful firewalling
    %primitives to understand whether traffic for this connection has been seen
    %before. With this new functionality, BPF programs can integrate more
    %closely with the networking stack's understanding of the traffic transiting
    %the kernel.

    %provide safe access to sockets from BPF packet handling hooks. The BPF verifier
    %is extended to gain understanding of reference-counted pointer types, which
    %allows BPF code paths to

\end{abstract}

\maketitle

\section{Keywords}

BPF, Linux, packet processing, sockets

\section{Introduction}

\begin{itemize}
\item BPF started with sockets, but only provided access to packets
\item Previous helper extensions allowed e.g. socket operation handling
\item Arbitrary socket lookup from socket/packet lookups allows more powerful
      BPF programs
\end{itemize}

\todo{Rewrite introduction}

% Over the past several years, BPF has steadily become more powerful in
% multiple ways: Through building more intelligence into the verifier which
% allows more complex programs to be loaded, and through extension of the API
% such as by adding new map types and new native BPF function calls. While
% BPF has its roots in applying filters at the socket layer, the ability to
% introspect the sockets relating to traffic being filtered has been limited.

%To build such awareness into a BPF helper, the verifier needs the ability to
%track the safety of the calls, including appropriate access into socket
%pointer types, and reference counting upon the underlying socket. This paper
%walks through extensions to the verifier to perform tracking of references in
%a BPF program. This allows BPF developers to extend the API with functions
%that allocate and release resources within the execution lifetime of a BPF
%program, and the verifier will validate that the resource acquisition and
%release is balanced across all execution paths of the BPF programs.

%Using this new reference tracking ability in the verifier, we add socket lookup
%and release function calls to the BPF API, allowing BPF programs to safely find
%a socket and build logic upon the presence or attributes of a socket. This can
%be used to load-balance traffic based on the presence of a listening
%application, or to implement stateful firewalling primitives to understand
%whether traffic for this connection has been seen before. With this new
%functionality, BPF programs can integrate more closely with the networking
%stack's understanding of the traffic transiting the kernel.

\todo{Desire to target logic based upon socket existence/attributes at packet
layers}

\section{Extending the Verifier}

%To validate access around sockets, the verifier must have knowledge of the
%socket type to validate that access is within bounds and is safe with respect
%to read or write.

The BPF verifier must ensure that all BPF programs that are loaded into the
kernel are safe to execute. For accessing socket pointers, this requires
validation that memory accesses are within the bounds of the memory allocated
for the structure, and will not make any modifications to the socket fields
that would be incompatible with the core socket handling logic. This is
provided through extension of existing functionality which provides bounds
verification and offset conversion for programs operating on CGroups hooks.
Additionally, the underlying socket memory must remain associated with the
socket for the duration of accesses from the BPF program, which may be
guaranteed by taking a reference on sockets for the duration of their use.

\subsection{Pointers to Sockets}

Earlier work~\cite{bpf-sock} introduced a \verb+bpf_sock+ structure to the BPF
API and implemented bounds and access type checking, along with offset
rewriting for converting the access of socket attributes from BPF instructions
into equivalent accesses of the underlying kernel types. This work was
introduced with new BPF hook points which provide the socket structure as the
context of the BPF program.

To allow BPF programs to retrieve and access a socket pointer, the verifier
must be made aware of when a register contains a pointer of this type, and also
how to validate pointer access. The existing verifier logic handles context
pointers in a generic manner via \verb+bpf_verifier_ops+. However, the packet
event hook points targeted in this paper already contain a context pointer of a
type that is different from \verb+bpf_sock+, so some additional logic was
required in the verifier to understand this pointer type. The patch series
associated with this paper introduced a new pointer type specific to sockets,
and linked its validation directly to the aforementioned socket verification
functions without using the verifier operations abstraction. Write operations
into the socket are rejected in the current implementation.

With the verifier now aware of socket pointer types, this could be built upon
to implicitly associate socket pointers with references to kernel resources.

\subsection{Reference tracking}

As with many other structure types, Linux sockets are reference-counted to
track usage and handle memory release. In general to ensure the safety of
socket for the duration of access in BPF programs, this reference count needs
to be incremented when sockets are referenced from BPF programs. Two options
were considered for how to guarantee this: Implicit reference tracking and
explicit reference tracking.

\subsubsection{Implicit reference tracking.}

This calls for the BPF infrastructure to handle reference tracking, and to hide
this detail from the BPF API. Whenever a socket lookup function is called, a
reference is taken and the socket is added to a reference list. At the end of
the BPF program execution, the core code could walk this reference list and
release each reference. However, collecting references and releasing them at
the end of the BPF program invocation has the unfortunate overhead that even
the execution of BPF programs that do not make use of the new socket lookup
helpers would need to pay the cost of checking the list of references to free.
When dealing with sufficiently demanding use cases such as those which use XDP,
even the few instructions required to implement such a check may have a
measurable impact on performance.

\subsubsection{Explicit reference tracking.}

This calls for reference acquisition and release semantics to be built into the
API. If references must be taken, then requiring BPF program writers to
explicitly handle these references ensures that the program writer must
understand the potential impact this may have on the operation of the program
(including atomic operation cost), and it also ensures that the cost is only
borne by programs that make use of this feature.

\subsubsection{Implementation.}

Based upon the tradeoffs described above, explicit reference tracking was
chosen. From an implementation perspective, this requires tracking pointer
values through each conditional path in the program, and rejecting the load of
programs that fail to "balance" resource acquisition (lookup) with release.
This implementation works as follows.

Calls to helper functions which acquire references to resources are annotated
in the verifier to associate the acquisition and release functions. When the
instruction that acquires a resource is processed, a resource identifier is
allocated for this resource. This identifier is kept in a list in the verifier
state, and it is also associated with the register which receives the resulting
resource pointer. When processing a corresponding release helper call, if a
parameter to the function contains a register that is associated with a
resource identifier, then the identifier is removed from the verifier state. If
any paths in the program reach the final (\verb+BPF_EXIT+) instruction and the
verifier state contains any resource identifier association, then the program
is considered unsafe as it has leaked the resource.

Some instruction calls are restricted while holding a resource reference to
avoid leaking references, for example the \verb+bpf_tail_call+ helper function.

\subsection{Runtime protection}

The verifier is tasked with ensuring that all BPF program execution flow is
safe to the extents where it interacts with the BPF API. In general, BPF
programs, once verified and triggered by event hooks, will execute from the
first instruction through the final instruction for a particular path through
the program. Therefore, after execution of a BPF program containing resource
acquire and release functions which are verified by the logic above, the
underlying resources should not leak.

One exception to this is when classic BPF \verb+LD_ABS+ or \verb+LD_IND+
instructions are used in the BPF program. These instructions were previously
used to provide direct packet access, with the semantics that if the access
offset exceeds the length of the packet, the BPF program would be terminated
prior to the final state. Without mitigation, this could lead to leaking of
references. Newer BPF APIs provide better alternatives for packet access, so we
disallow the use of these instructions while holding a socket reference.

\section{Extending the BPF API}

To implement explicit reference tracking, the API required at least two
functions to handle referencing of sockets from the BPF API: A socket lookup
function and a socket release function. This section describes the API
considerations for each of these.

\subsection{API definition}

The following considerations were made for implementing the lookup helpers. The
consideration of each of these items reflects the definition of the lookup
helpers in Listing~\ref{lst:helper-api}.

\begin{itemize}
    \item The Linux stack may contain multiple network namespaces, where the
          BPF program may operate attached to a device in one network namespace
          while the desired socket may exist in another network namespace.
    \item BPF programmers may wish to discover sockets that are not directly
          related to the BPF program context (packet, socket, etc).
    \item Current use cases are focused around UDP and TCP sockets, however
          there may be a desire to extend this in future.
\end{itemize}

\todo{Span this listing across two columns}
\begin{lstlisting}[caption={BPF API helper functions for socket lookup},language=c,label=lst:helper-api,breaklines=true]
struct bpf_sock *bpf_sk_lookup_tcp(void *ctx, struct bpf_sock_tuple *tuple, u32 tuple_size,
                                   u32 netns, u64 flags)
struct bpf_sock *bpf_sk_lookup_udp(void *ctx, struct bpf_sock_tuple *tuple, u32 tuple_size,
                                   u32 netns, u64 flags)
int bpf_sk_release(struct bpf_sock *sk)
\end{lstlisting}

\subsubsection{Namespacing.}

Linux network devices are associated with a particular network namespace which
provides logical network stack separation. Sockets are also inherently
associated with an application that operates in a particular network namespace.
The simplest path for handling namespacing from the lookup handlers would be to
allow BPF programs operating on a device within a network namespace to only
find sockets in the same network namespace. By passing the BPF program context
into the socket lookup helper, the implementation can derive the source network
namespace from this context. Beyond this, with the growing popularity of
containers it may be desirable for a network orchestration tool to make use of
socket properties inside container network namespaces to influence decisions
made on packets outside the container---for instance, on packets being received
into a node prior to the packet passing into the container network namespace,
or if filtering is implemented on a veth pair outside of a container. By
allowing the netns ID to be specified, this use case can be supported as well.

\subsubsection{Tuple definition.}

A common use case for the socket lookup functions would call them from a packet
context, where the \verb+__sk_buff+ has the full 5-tuple information available
and easily accessible. One could imagine a simpler helper function that allows
the BPF program writer to provide the \verb+__sk_buff+ to the helper function,
then the helper function would look up the socket based upon the packet
metadata. This would however limit the potential uses for such a helper. Two
cases that would be more limited with this model are usage from XDP, or if the
BPF program implements a form of network address translation. From XDP, if the
raw \verb+xdp_md+ context is provided to such a helper, then the helper would
need to parse the packet to understand the 5-tuple. This would duplicate the
standard stack paths which are executed after XDP where the \verb+sk_buff+ is
built, but this information would be subsequently thrown away upon return of
the helper call. Adding to this, if the program performs network address
translation between the local application and the remote destination, then the
local stack may not contain a socket associated with the tuple that is directly
in the packet. While these cases could be worked around with a simpler helper,
it was deemed more powerful and generic to allow the BPF program writer to
provide the tuple for lookup.

\subsubsection{Extensibility.}

New kernel APIs that have any scope for extension should contain a flags
argument. For the socket lookup helpers, a few ideas had been considered as
potential future alternatives to the existing behaviour. The lookup functions
follow the standard socket lookup paths in Linux which have predetermined
methods for selecting a socket when the application uses \verb+SO_REUSEPORT+.
Some discussion on the mailinglist proposed allowing BPF program writers to
influence the socket selection mechanism~\cite{netdev-sk-select}.

\subsubsection{Result.}

Another aspect of extensibility is the ability to lookup sockets which are not
TCP or UDP. The initial RFC of this patch series proposed a single lookup
function which would choose the L4 protocol based upon a field in the
tuple~\cite{sk-lookup-rfc}, however this would make it more difficult for BPF
program writers to detect the support for different protocol types at compile
time. There is not a significant number of supported L4 socket types in the
Linux stack today, and the number is not expected to increase drastically, so
it was considered simpler and easier to split out each L4 socket lookup
function into an independent helper call.

\subsection{Optimizations}

Multiple optimizations were proposed during development of this feature to
reduce the runtime overhead of using the socket lookup helpers. Two such
optimizations are described below: Skipping reference counting when unnecessary
to save on atomic instructions, and allowing the use of direct packet pointers
for the tuple.

\subsubsection{Avoid reference counting when unnecessary.}

For some socket types, such as UDP sockets, or TCP listen sockets, memory
access safety can be achieved with minimal implementation: The destruction of
such sockets is already governed by standard RCU rules, meaning that while the
RCU lock is held, they can be safely referenced; Once the RCU grace period is
reached, the memory may be freed and references to the socket are no longer
safe to use. Because BPF programs run under the RCU lock, this allows simple,
safe access of properties of these sockets without requiring additional
mechanisms to ensure the safety of the socket pointers.

When it comes to TCP sockets that are not governed by RCU, the networking stack
uses reference counting to manage socket memory instead. To ensure that the
socket remains safe to access for a period of BPF program execution, the BPF
program must also take a reference on the socket. This also implies that such
references must be subsequently released to prevent leaking kernel resources.

\subsubsection{Allow lookup using direct packet pointers.}

The tuple structure is defined in such a way that if the IPv4 or IPv6 packet is
immediately followed by the TCP or UDP header without IP options in between,
then a pointer directly to the L3 addresses may be passed to the lookup
function, allowing the helper function to directly pull the addresses and ports
from the packet buffer rather than requiring the BPF program to first extract
these onto the stack.

\subsection{Future work}

The implementation introduced in the 4.20 release cycle includes support for
looking up TCP and UDP sockets for IPv4 and IPv6 traffic, using the
\verb+SCHED_CLS+, \verb+SCHED_ACT+ and \verb+SK_SKB+ hook points. A proposed
patchset extends this to allow the same helpers to be used from the \verb+XDP+
hook points~\cite{sk-lookup-xdp}.

\todo{Associate custom metadata with socket?}

\section{Use cases}

\subsection{Stateful Firewalling}

\subsection{Load-balancing}

\section{Conclusion}

This paper describes the contribution of new BPF verifier functionality in the
Linux kernel to track references to kernel resources and uses these to provide
access to socket introspection capabilities from packet handling hooks. The
introduction of reference tracking logic into the verifier provides
useful base infrastructure for supporting acquire and release semantics for
kernel resources which may prove useful for other helpers in future, and the
socket lookup API makes the packet hooks more powerful to support use cases
such as stateful firewalling or load-balancing based on the sockets that are
open on a system.

\section{Acknowledgments}

The core BPF maintainers provided great guidance and review of the patches, in
particular Alexei Starovoitov, Daniel Borkmann and Martin KaFai Lau.

\bibliographystyle{plainnat}
\bibliography{paper}

\end{document}

### 2018-10-27 paper/paper.tex
paper: Flesh out general structure and some text

@@
     capabilities of the Linux kernel's BPF packet hooks. To provide safety
-    guarantees around the access of the socket, the verifier is extended to
+    guarantees around the access of the socket, the verifier was extended to
     track allocation and release of kernel resources. Using this mechanism, we
     introduce socket lookup and release helpers into the BPF API.
-
-    %These new primitives open up new possibilities for various use cases
-    %including stateful firewalling and load balancing.
 
@@
 
-\begin{itemize}
-\item BPF started with sockets, but only provided access to packets
-\item Previous helper extensions allowed e.g. socket operation handling
-\item Arbitrary socket lookup from socket/packet lookups allows more powerful
-      BPF programs
-\end{itemize}
-
-\todo{Rewrite introduction}
-
 % Over the past several years, BPF has steadily become more powerful in
@@
 
-%To build such awareness into a BPF helper, the verifier needs the ability to
-%track the safety of the calls, including appropriate access into socket
-%pointer types, and reference counting upon the underlying socket. This paper
-%walks through extensions to the verifier to perform tracking of references in
-%a BPF program. This allows BPF developers to extend the API with functions
-%that allocate and release resources within the execution lifetime of a BPF
-%program, and the verifier will validate that the resource acquisition and
-%release is balanced across all execution paths of the BPF programs.
+To build such awareness into a BPF helper, the verifier needs the ability to
+track the safety of the calls, including appropriate reference counting upon
+the underlying socket. This talk walks through extensions to the verifier to
+perform tracking of references in a BPF program. This allows BPF developers to
+extend the UAPI with functions that allocate and release resources within the
+execution lifetime of a BPF program, and the verifier will validate that the
+resources are released exactly once prior to program completion.
 
-%Using this new reference tracking ability in the verifier, we add socket lookup
-%and release function calls to the BPF API, allowing BPF programs to safely find
-%a socket and build logic upon the presence or attributes of a socket. This can
-%be used to load-balance traffic based on the presence of a listening
-%application, or to implement stateful firewalling primitives to understand
-%whether traffic for this connection has been seen before. With this new
-%functionality, BPF programs can integrate more closely with the networking
-%stack's understanding of the traffic transiting the kernel.
+Using this new reference tracking ability in the verifier, we add socket lookup
+and release function calls to the BPF API, allowing BPF programs to safely find
+a socket and build logic upon the presence or attributes of a socket. This can
+be used to load-balance traffic based on the presence of a listening
+application, or to implement stateful firewalling primitives to understand
+whether traffic for this connection has been seen before. With this new
+functionality, BPF programs can integrate more closely with the networking
+stack's understanding of the traffic transiting the kernel.
 
-\todo{Desire to target logic based upon socket existence/attributes at packet
-layers}
+This safety is provided by tracking pointer values through each conditional
+path in the program, and rejecting the load of programs that fail to "balance"
+resource acquisition (lookup) with release.
+
+\section{Background}
+
+The BPF verifier must ensure that all BPF programs that are loaded into the
+kernel are safe to execute. When accessing kernel resources, this includes
+providing guarantees around the availability of the underlying resource for the
+duration of the BPF program. For some socket types, such as UDP sockets, or TCP
+listen sockets, this can be achieved with minimal implementation: The
+destruction of such sockets is already governed by standard RCU rules, meaning
+that while the RCU lock is held, they can be safely referenced; Once the RCU
+grace period is reached, the memory may be freed and references to the socket
+are no longer safe to use. Given that BPF programs run under the RCU lock, this
+allows simple, safe access of properties of these sockets without requiring
+additional mechanisms to ensure the safety of the socket pointers.
+
+When it comes to TCP sockets that are not governed by RCU, the standard
+refcounting model is used instead. To ensure that the socket remains safe to
+access for a period of BPF program execution, the BPF program must also take a
+reference on the socket. Furthermore, if references must be taken, then they
+also must be released.
 
@@
 
-%To validate access around sockets, the verifier must have knowledge of the
-%socket type to validate that access is within bounds and is safe with respect
-%to read or write.
-
-The BPF verifier must ensure that all BPF programs that are loaded into the
-kernel are safe to execute. For accessing socket pointers, this requires
-validation that memory accesses are within the bounds of the memory allocated
-for the structure, and will not make any modifications to the socket fields
-that would be incompatible with the core socket handling logic. This is
-provided through extension of existing functionality which provides bounds
-verification and offset conversion for programs operating on CGroups hooks.
-Additionally, the underlying socket memory must remain associated with the
-socket for the duration of accesses from the BPF program, which may be
-guaranteed by taking a reference on sockets for the duration of their use.
-
 \subsection{Pointers to Sockets}
-
-Earlier work~\cite{bpf-sock} introduced a \verb+bpf_sock+ structure to the BPF
-API and implemented bounds and access type checking, along with offset
-rewriting for converting the access of socket attributes from BPF instructions
-into equivalent accesses of the underlying kernel types. This work was
-introduced with new BPF hook points which provide the socket structure as the
-context of the BPF program.
-
-To allow BPF programs to retrieve and access a socket pointer, the verifier
-must be made aware of when a register contains a pointer of this type, and also
-how to validate pointer access. The existing verifier logic handles context
-pointers in a generic manner via \verb+bpf_verifier_ops+. However, the packet
-event hook points targeted in this paper already contain a context pointer of a
-type that is different from \verb+bpf_sock+, so some additional logic was
-required in the verifier to understand this pointer type. The patch series
-associated with this paper introduced a new pointer type specific to sockets,
-and linked its validation directly to the aforementioned socket verification
-functions without using the verifier operations abstraction. Write operations
-into the socket are rejected in the current implementation.
-
-With the verifier now aware of socket pointer types, this could be built upon
-to implicitly associate socket pointers with references to kernel resources.
 
@@
 
-As with many other structure types, Linux sockets are reference-counted to
-track usage and handle memory release. In general to ensure the safety of
-socket for the duration of access in BPF programs, this reference count needs
-to be incremented when sockets are referenced from BPF programs. Two options
-were considered for how to guarantee this: Implicit reference tracking and
-explicit reference tracking.
+Alternatives:
+\begin{itemize}
+\item RCU-protect other socket types
+\item Collect references during program execution and release afterwards
+\item Build acquire/release semantics into the API
+\end{itemize}
 
-\subsubsection{Implicit reference tracking.}
-
-This calls for the BPF infrastructure to handle reference tracking, and to hide
-this detail from the BPF API. Whenever a socket lookup function is called, a
-reference is taken and the socket is added to a reference list. At the end of
-the BPF program execution, the core code could walk this reference list and
-release each reference. However, collecting references and releasing them at
-the end of the BPF program invocation has the unfortunate overhead that even
-the execution of BPF programs that do not make use of the new socket lookup
-helpers would need to pay the cost of checking the list of references to free.
-When dealing with sufficiently demanding use cases such as those which use XDP,
-even the few instructions required to implement such a check may have a
-measurable impact on performance.
-
-\subsubsection{Explicit reference tracking.}
-
-This calls for reference acquisition and release semantics to be built into the
-API. If references must be taken, then requiring BPF program writers to
-explicitly handle these references ensures that the program writer must
-understand the potential impact this may have on the operation of the program
-(including atomic operation cost), and it also ensures that the cost is only
-borne by programs that make use of this feature.
-
-\subsubsection{Implementation.}
-
-Based upon the tradeoffs described above, explicit reference tracking was
-chosen. From an implementation perspective, this requires tracking pointer
-values through each conditional path in the program, and rejecting the load of
-programs that fail to "balance" resource acquisition (lookup) with release.
-This implementation works as follows.
+Collecting references and releasing afterwards has the unfortunate overhead
+that even the execution of BPF programs that do not make use of the new socket
+lookup helpers would need to pay the cost of checking the list of references
+to free. When dealing with sufficiently demanding use cases such as those which
+use XDP, even the few instructions required to implement such a check may have
+a measurable impact on performance.
 
@@
 in the verifier to associate the acquisition and release functions. When the
-instruction that acquires a resource is processed, a resource identifier is
-allocated for this resource. This identifier is kept in a list in the verifier
-state, and it is also associated with the register which receives the resulting
-resource pointer. When processing a corresponding release helper call, if a
-parameter to the function contains a register that is associated with a
-resource identifier, then the identifier is removed from the verifier state. If
-any paths in the program reach the final (\verb+BPF_EXIT+) instruction and the
-verifier state contains any resource identifier association, then the program
-is considered unsafe as it has leaked the resource.
-
-Some instruction calls are restricted while holding a resource reference to
-avoid leaking references, for example the \verb+bpf_tail_call+ helper function.
+instruction that acquires a resource is processed, the verifier state is
+updated to include a resource identifier for the resulting value, and the
+register's value is associated with this identifier. When processing a
+corresponding release helper call, if a parameter to the function contains a
+resource identifier, then the corresponding identifire is removed from the
+verifier state. If any paths in the program reach the final (\verb+BPF_EXIT+)
+instruction and the verifier state still contains resource identifiers, then
+the program is considered unsafe as it has leaked the resource.
 
@@
 
-The verifier is tasked with ensuring that all BPF program execution flow is
-safe to the extents where it interacts with the BPF API. In general, BPF
-programs, once verified and triggered by event hooks, will execute from the
-first instruction through the final instruction for a particular path through
-the program. Therefore, after execution of a BPF program containing resource
-acquire and release functions which are verified by the logic above, the
-underlying resources should not leak.
-
-One exception to this is when classic BPF \verb+LD_ABS+ or \verb+LD_IND+
-instructions are used in the BPF program. These instructions were previously
-used to provide direct packet access, with the semantics that if the access
-offset exceeds the length of the packet, the BPF program would be terminated
-prior to the final state. Without mitigation, this could lead to leaking of
-references. Newer BPF APIs provide better alternatives for packet access, so we
-disallow the use of these instructions while holding a socket reference.
+\todo{LD\_ABS, LD\_IND}
 
 \section{Extending the BPF API}
-
-To implement explicit reference tracking, the API required at least two
-functions to handle referencing of sockets from the BPF API: A socket lookup
-function and a socket release function. This section describes the API
-considerations for each of these.
-
-\subsection{API definition}
-
-The following considerations were made for implementing the lookup helpers. The
-consideration of each of these items reflects the definition of the lookup
-helpers in Listing~\ref{lst:helper-api}.
-
-\begin{itemize}
-    \item The Linux stack may contain multiple network namespaces, where the
-          BPF program may operate attached to a device in one network namespace
-          while the desired socket may exist in another network namespace.
-    \item BPF programmers may wish to discover sockets that are not directly
-          related to the BPF program context (packet, socket, etc).
-    \item Current use cases are focused around UDP and TCP sockets, however
-          there may be a desire to extend this in future.
-\end{itemize}
-
-\todo{Span this listing across two columns}
-\begin{lstlisting}[caption={BPF API helper functions for socket lookup},language=c,label=lst:helper-api,breaklines=true]
-struct bpf_sock *bpf_sk_lookup_tcp(void *ctx, struct bpf_sock_tuple *tuple, u32 tuple_size,
-                                   u32 netns, u64 flags)
-struct bpf_sock *bpf_sk_lookup_udp(void *ctx, struct bpf_sock_tuple *tuple, u32 tuple_size,
-                                   u32 netns, u64 flags)
-int bpf_sk_release(struct bpf_sock *sk)
-\end{lstlisting}
-
-\subsubsection{Namespacing.}
-
-Linux network devices are associated with a particular network namespace which
-provides logical network stack separation. Sockets are also inherently
-associated with an application that operates in a particular network namespace.
-The simplest path for handling namespacing from the lookup handlers would be to
-allow BPF programs operating on a device within a network namespace to only
-find sockets in the same network namespace. By passing the BPF program context
-into the socket lookup helper, the implementation can derive the source network
-namespace from this context. Beyond this, with the growing popularity of
-containers it may be desirable for a network orchestration tool to make use of
-socket properties inside container network namespaces to influence decisions
-made on packets outside the container---for instance, on packets being received
-into a node prior to the packet passing into the container network namespace,
-or if filtering is implemented on a veth pair outside of a container. By
-allowing the netns ID to be specified, this use case can be supported as well.
-
-\subsubsection{Tuple definition.}
-
-A common use case for the socket lookup functions would call them from a packet
-context, where the \verb+__sk_buff+ has the full 5-tuple information available
-and easily accessible. One could imagine a simpler helper function that allows
-the BPF program writer to provide the \verb+__sk_buff+ to the helper function,
-then the helper function would look up the socket based upon the packet
-metadata. This would however limit the potential uses for such a helper. Two
-cases that would be more limited with this model are usage from XDP, or if the
-BPF program implements a form of network address translation. From XDP, if the
-raw \verb+xdp_md+ context is provided to such a helper, then the helper would
-need to parse the packet to understand the 5-tuple. This would duplicate the
-standard stack paths which are executed after XDP where the \verb+sk_buff+ is
-built, but this information would be subsequently thrown away upon return of
-the helper call. Adding to this, if the program performs network address
-translation between the local application and the remote destination, then the
-local stack may not contain a socket associated with the tuple that is directly
-in the packet. While these cases could be worked around with a simpler helper,
-it was deemed more powerful and generic to allow the BPF program writer to
-provide the tuple for lookup.
-
-\subsubsection{Extensibility.}
-
-New kernel APIs that have any scope for extension should contain a flags
-argument. For the socket lookup helpers, a few ideas had been considered as
-potential future alternatives to the existing behaviour. The lookup functions
-follow the standard socket lookup paths in Linux which have predetermined
-methods for selecting a socket when the application uses \verb+SO_REUSEPORT+.
-Some discussion on the mailinglist proposed allowing BPF program writers to
-influence the socket selection mechanism~\cite{netdev-sk-select}.
-
-\subsubsection{Result.}
-
-Another aspect of extensibility is the ability to lookup sockets which are not
-TCP or UDP. The initial RFC of this patch series proposed a single lookup
-function which would choose the L4 protocol based upon a field in the
-tuple~\cite{sk-lookup-rfc}, however this would make it more difficult for BPF
-program writers to detect the support for different protocol types at compile
-time. There is not a significant number of supported L4 socket types in the
-Linux stack today, and the number is not expected to increase drastically, so
-it was considered simpler and easier to split out each L4 socket lookup
-function into an independent helper call.
-
-\subsection{Optimizations}
-
-Multiple optimizations were proposed during development of this feature to
-reduce the runtime overhead of using the socket lookup helpers. Two such
-optimizations are described below: Skipping reference counting when unnecessary
-to save on atomic instructions, and allowing the use of direct packet pointers
-for the tuple.
-
-\subsubsection{Avoid reference counting when unnecessary.}
-
-For some socket types, such as UDP sockets, or TCP listen sockets, memory
-access safety can be achieved with minimal implementation: The destruction of
-such sockets is already governed by standard RCU rules, meaning that while the
-RCU lock is held, they can be safely referenced; Once the RCU grace period is
-reached, the memory may be freed and references to the socket are no longer
-safe to use. Because BPF programs run under the RCU lock, this allows simple,
-safe access of properties of these sockets without requiring additional
-mechanisms to ensure the safety of the socket pointers.
-
-When it comes to TCP sockets that are not governed by RCU, the networking stack
-uses reference counting to manage socket memory instead. To ensure that the
-socket remains safe to access for a period of BPF program execution, the BPF
-program must also take a reference on the socket. This also implies that such
-references must be subsequently released to prevent leaking kernel resources.
-
-\subsubsection{Allow lookup using direct packet pointers.}
-
-The tuple structure is defined in such a way that if the IPv4 or IPv6 packet is
-immediately followed by the TCP or UDP header without IP options in between,
-then a pointer directly to the L3 addresses may be passed to the lookup
-function, allowing the helper function to directly pull the addresses and ports
-from the packet buffer rather than requiring the BPF program to first extract
-these onto the stack.
-
-\subsection{Future work}
-
-The implementation introduced in the 4.20 release cycle includes support for
-looking up TCP and UDP sockets for IPv4 and IPv6 traffic, using the
-\verb+SCHED_CLS+, \verb+SCHED_ACT+ and \verb+SK_SKB+ hook points. A proposed
-patchset extends this to allow the same helpers to be used from the \verb+XDP+
-hook points~\cite{sk-lookup-xdp}.
-
-\todo{Associate custom metadata with socket?}
 
@@
 
-This paper describes the contribution of new BPF verifier functionality in the
-Linux kernel to track references to kernel resources and uses these to provide
-access to socket introspection capabilities from packet handling hooks. The
-introduction of reference tracking logic into the verifier provides
-useful base infrastructure for supporting acquire and release semantics for
-kernel resources which may prove useful for other helpers in future, and the
-socket lookup API makes the packet hooks more powerful to support use cases
-such as stateful firewalling or load-balancing based on the sockets that are
-open on a system.
+This paper describes the contribution of two extensions to the Linux kernel BPF
+implementation, firstly the introduction of reference tracking logic into the
+verifier, which provides useful base infrastructure for supporting acquire and
+release semantics for kernel resources, and secondly the socket lookup API
+which may be now used from packet paths to allow BPF writers to build logic
+based upon introspection of kernel sockets.
 
@@
 
-The core BPF maintainers provided great guidance and review of the patches, in
-particular Alexei Starovoitov, Daniel Borkmann and Martin KaFai Lau.
+The core BPF maintainers for their guidance and review of the patches.
 

### 2018-10-27 paper/paper.tex
paper: Add stateful firewalling

--- 2018-10-27 paper/paper.tex
@@
 
-    This paper presents the changes necessary to extend socket introspection
-    capabilities of the Linux kernel's BPF packet hooks. To provide safety
-    guarantees around the access of the socket, the verifier is extended to
-    track allocation and release of kernel resources. Using this mechanism, we
-    introduce socket lookup and release helpers into the BPF API.
+    To build such awareness into a BPF helper, the verifier needs the ability
+    to track the safety of the calls, including appropriate reference counting
+    upon the underlying socket. This paper describes extensions to the verifier
+    to perform tracking of references in a BPF program. This allows BPF
+    developers to extend the UAPI with functions that allocate and release
+    resources within the execution lifetime of a BPF program, and the verifier
+    will validate that the resources are released exactly once prior to program
+    completion.
 
-    %These new primitives open up new possibilities for various use cases
-    %including stateful firewalling and load balancing.
-
-    %To provide access to socket state from a BPF helper call, the verifier
-    %needs the ability to track the safety of the calls, including appropriate
-    %reference counting upon the underlying socket for the duration of the
-    %access.
-
-    %With such an ability ability in the verifier, we add socket lookup and
-    %release function calls to the BPF API, allowing BPF programs to safely find
-    %a socket and build logic upon the presence or attributes of a socket.
-    %Applications for this functionality range from load-balancing traffic based
-    %on the presence of a listening application to building stateful firewalling
-    %primitives to understand whether traffic for this connection has been seen
-    %before. With this new functionality, BPF programs can integrate more
-    %closely with the networking stack's understanding of the traffic transiting
-    %the kernel.
-
-    %provide safe access to sockets from BPF packet handling hooks. The BPF verifier
-    %is extended to gain understanding of reference-counted pointer types, which
-    %allows BPF code paths to
+    Using this new reference tracking ability in the verifier, we add socket
+    lookup and release function calls to the BPF API, allowing BPF programs to
+    safely find a socket and build logic upon the presence or attributes of a
+    socket. This can be used to load-balance traffic based on the presence of a
+    listening application, or to implement stateful firewalling primitives to
+    understand whether traffic for this connection has been seen before. With
+    this new functionality, BPF programs can integrate more closely with the
+    networking stack's understanding of the traffic transiting the kernel.
 
@@
 \item Previous helper extensions allowed e.g. socket operation handling
-\item Arbitrary socket lookup from socket/packet lookups allows more powerful
-      BPF programs
+\item Arbitrary socket lookup from socket/packet hooks allows more powerful BPF
+      programs
+\item Use cases revolve around building logic in socket/packet layer hooks that
+      change the behaviour based on the existence and attributes of a socket.
 \end{itemize}
-
-\todo{Rewrite introduction}
-
-% Over the past several years, BPF has steadily become more powerful in
-% multiple ways: Through building more intelligence into the verifier which
-% allows more complex programs to be loaded, and through extension of the API
-% such as by adding new map types and new native BPF function calls. While
-% BPF has its roots in applying filters at the socket layer, the ability to
-% introspect the sockets relating to traffic being filtered has been limited.
-
-%To build such awareness into a BPF helper, the verifier needs the ability to
-%track the safety of the calls, including appropriate access into socket
-%pointer types, and reference counting upon the underlying socket. This paper
-%walks through extensions to the verifier to perform tracking of references in
-%a BPF program. This allows BPF developers to extend the API with functions
-%that allocate and release resources within the execution lifetime of a BPF
-%program, and the verifier will validate that the resource acquisition and
-%release is balanced across all execution paths of the BPF programs.
-
-%Using this new reference tracking ability in the verifier, we add socket lookup
-%and release function calls to the BPF API, allowing BPF programs to safely find
-%a socket and build logic upon the presence or attributes of a socket. This can
-%be used to load-balance traffic based on the presence of a listening
-%application, or to implement stateful firewalling primitives to understand
-%whether traffic for this connection has been seen before. With this new
-%functionality, BPF programs can integrate more closely with the networking
-%stack's understanding of the traffic transiting the kernel.
-
-\todo{Desire to target logic based upon socket existence/attributes at packet
-layers}
 
@@
 that would be incompatible with the core socket handling logic. This is
-provided through extension of existing functionality which provides bounds
+provided through extension of existing functionality which performs bounds
 verification and offset conversion for programs operating on CGroups hooks.
-Additionally, the underlying socket memory must remain associated with the
-socket for the duration of accesses from the BPF program, which may be
-guaranteed by taking a reference on sockets for the duration of their use.
+Additionally, the underlying memory must remain associated with the socket for
+the duration of accesses from the BPF program, which may be guaranteed by
+taking a reference on sockets for the duration of their use. The following
+sections explain how this is implemented through the introduction of a new
+pointer type in the verifier, and through the extension of the verifier to
+recognise assembly instructions that represent reference acquisition and
+release.
 
@@
 
-As with many other structure types, Linux sockets are reference-counted to
-track usage and handle memory release. In general to ensure the safety of
-socket for the duration of access in BPF programs, this reference count needs
-to be incremented when sockets are referenced from BPF programs. Two options
-were considered for how to guarantee this: Implicit reference tracking and
-explicit reference tracking.
+Sockets in Linux are reference-counted to track usage and handle memory
+release. In general to ensure the safety of socket for the duration of access
+in BPF programs, this reference count needs to be incremented when sockets are
+referenced from BPF programs. Two options were considered for how to guarantee
+this: Implicit reference tracking and explicit reference tracking.
 
@@
 
+Stateful firewalling is distinguished from stateless firewalling in that it
+provides the ability to associate two directions of a connection together and
+apply a filtering policy based upon the direction of the connection. This
+commonly involves using a connection tracker to reconstruct the connection
+state of each endpoint of the connection without specific knowledge from those
+endpoints~\cite{ovs-ct,conntrack}. In cases where the Linux stack acts as a
+firewall on a node in the network that is separate from both endpoints, In many
+cases this model is used when filtering traffic where one endpoint is colocated
+with the Linux stack that performs the firewalling, which already contains
+information about the local end of the connection in a slightly.
+
 \subsection{Load-balancing}

### 2018-10-27 paper/paper.tex
Import LPC template

--- 2018-10-27 paper/paper.tex
@@
 \pdfinfo{
-/Title (Building socket-aware BPF programs)
-/Author (Joe Stringer)}
-
-\title{Building socket-aware BPF programs}
-\author{Joe Stringer}
-\affiliation{%
-      \institution{Cilium.io}}
-\email{joe@cilium.io}
+/Title (Title)
+/Author (Author)}
+\title{Title}
+\author{Author\\
+Affiliation\\
+Email\\
+}
 
@@
 
-\usepackage{xcolor}
-\newcommand\todo[1]{\textcolor{red}{#1}}
+\begin{document}
 
-\begin{document}
+\maketitle
 
@@
 
-    Over the past several years, BPF has steadily become more powerful in
-    multiple ways: Through building more intelligence into the verifier which
-    allows more complex programs to be loaded, and through extension of the API
-    such as by adding new map types and new native BPF function calls. While
-    BPF has its roots in applying filters at the socket layer, the ability to
-    introspect the sockets relating to traffic being filtered has been limited.
-
-    This paper presents the changes necessary to extend socket introspection
-    capabilities of the Linux kernel's BPF packet hooks. To provide safety
-    guarantees around the access of the socket, the verifier was extended to
-    track allocation and release of kernel resources. Using this mechanism, we
-    introduce socket lookup and release helpers into the BPF API.
-
-    %To provide access to socket state from a BPF helper call, the verifier
-    %needs the ability to track the safety of the calls, including appropriate
-    %reference counting upon the underlying socket for the duration of the
-    %access.
-
-    %With such an ability ability in the verifier, we add socket lookup and
-    %release function calls to the BPF API, allowing BPF programs to safely find
-    %a socket and build logic upon the presence or attributes of a socket.
-    %Applications for this functionality range from load-balancing traffic based
-    %on the presence of a listening application to building stateful firewalling
-    %primitives to understand whether traffic for this connection has been seen
-    %before. With this new functionality, BPF programs can integrate more
-    %closely with the networking stack's understanding of the traffic transiting
-    %the kernel.
-
-    %provide safe access to sockets from BPF packet handling hooks. The BPF verifier
-    %is extended to gain understanding of reference-counted pointer types, which
-    %allows BPF code paths to
-
 \end{abstract}
-
-\maketitle
 
@@
 
-BPF, Linux, packet processing, sockets
-
 \section{Introduction}
-
-% Over the past several years, BPF has steadily become more powerful in
-% multiple ways: Through building more intelligence into the verifier which
-% allows more complex programs to be loaded, and through extension of the API
-% such as by adding new map types and new native BPF function calls. While
-% BPF has its roots in applying filters at the socket layer, the ability to
-% introspect the sockets relating to traffic being filtered has been limited.
-
-To build such awareness into a BPF helper, the verifier needs the ability to
-track the safety of the calls, including appropriate reference counting upon
-the underlying socket. This talk walks through extensions to the verifier to
-perform tracking of references in a BPF program. This allows BPF developers to
-extend the UAPI with functions that allocate and release resources within the
-execution lifetime of a BPF program, and the verifier will validate that the
-resources are released exactly once prior to program completion.
-
-Using this new reference tracking ability in the verifier, we add socket lookup
-and release function calls to the BPF API, allowing BPF programs to safely find
-a socket and build logic upon the presence or attributes of a socket. This can
-be used to load-balance traffic based on the presence of a listening
-application, or to implement stateful firewalling primitives to understand
-whether traffic for this connection has been seen before. With this new
-functionality, BPF programs can integrate more closely with the networking
-stack's understanding of the traffic transiting the kernel.
-
-This safety is provided by tracking pointer values through each conditional
-path in the program, and rejecting the load of programs that fail to "balance"
-resource acquisition (lookup) with release.
-
-\section{Background}
-
-The BPF verifier must ensure that all BPF programs that are loaded into the
-kernel are safe to execute. When accessing kernel resources, this includes
-providing guarantees around the availability of the underlying resource for the
-duration of the BPF program. For some socket types, such as UDP sockets, or TCP
-listen sockets, this can be achieved with minimal implementation: The
-destruction of such sockets is already governed by standard RCU rules, meaning
-that while the RCU lock is held, they can be safely referenced; Once the RCU
-grace period is reached, the memory may be freed and references to the socket
-are no longer safe to use. Given that BPF programs run under the RCU lock, this
-allows simple, safe access of properties of these sockets without requiring
-additional mechanisms to ensure the safety of the socket pointers.
-
-When it comes to TCP sockets that are not governed by RCU, the standard
-refcounting model is used instead. To ensure that the socket remains safe to
-access for a period of BPF program execution, the BPF program must also take a
-reference on the socket. Furthermore, if references must be taken, then they
-also must be released.
-
-\section{Extending the Verifier}
-
-\subsection{Pointers to Sockets}
-
-\subsection{Reference tracking}
-
-Alternatives:
-\begin{itemize}
-\item RCU-protect other socket types
-\item Collect references during program execution and release afterwards
-\item Build acquire/release semantics into the API
-\end{itemize}
-
-Collecting references and releasing afterwards has the unfortunate overhead
-that even the execution of BPF programs that do not make use of the new socket
-lookup helpers would need to pay the cost of checking the list of references
-to free. When dealing with sufficiently demanding use cases such as those which
-use XDP, even the few instructions required to implement such a check may have
-a measurable impact on performance.
-
-Calls to helper functions which acquire references to resources are annotated
-in the verifier to associate the acquisition and release functions. When the
-instruction that acquires a resource is processed, the verifier state is
-updated to include a resource identifier for the resulting value, and the
-register's value is associated with this identifier. When processing a
-corresponding release helper call, if a parameter to the function contains a
-resource identifier, then the corresponding identifire is removed from the
-verifier state. If any paths in the program reach the final (\verb+BPF_EXIT+)
-instruction and the verifier state still contains resource identifiers, then
-the program is considered unsafe as it has leaked the resource.
-
-\subsection{Runtime protection}
-
-\todo{LD\_ABS, LD\_IND}
-
-\section{Extending the BPF API}
-
-\section{Use cases}
-
-\subsection{Stateful Firewalling}
-
-\subsection{Load-balancing}
 
@@
 
-This paper describes the contribution of two extensions to the Linux kernel BPF
-implementation, firstly the introduction of reference tracking logic into the
-verifier, which provides useful base infrastructure for supporting acquire and
-release semantics for kernel resources, and secondly the socket lookup API
-which may be now used from packet paths to allow BPF writers to build logic
-based upon introspection of kernel sockets.
-
 \section{Acknowledgments}
 
-The core BPF maintainers for their guidance and review of the patches.
-
 \bibliographystyle{plainnat}
-\bibliography{paper}
+%\bibliography{paper}
 

### 2018-10-27 paper/paper.tex
paper: Flesh out verifier, api definition sections

--- 2018-10-27 paper/paper.tex
@@
 
-    To build such awareness into a BPF helper, the verifier needs the ability
-    to track the safety of the calls, including appropriate reference counting
-    upon the underlying socket. This paper describes extensions to the verifier
-    to perform tracking of references in a BPF program. This allows BPF
-    developers to extend the UAPI with functions that allocate and release
-    resources within the execution lifetime of a BPF program, and the verifier
-    will validate that the resources are released exactly once prior to program
-    completion.
+    This paper presents the changes necessary to extend socket introspection
+    capabilities of the Linux kernel's BPF packet hooks. To provide safety
+    guarantees around the access of the socket, the verifier was extended to
+    track allocation and release of kernel resources. Using this mechanism, we
+    introduce socket lookup and release helpers into the BPF API.
 
-    Using this new reference tracking ability in the verifier, we add socket
-    lookup and release function calls to the BPF API, allowing BPF programs to
-    safely find a socket and build logic upon the presence or attributes of a
-    socket. This can be used to load-balance traffic based on the presence of a
-    listening application, or to implement stateful firewalling primitives to
-    understand whether traffic for this connection has been seen before. With
-    this new functionality, BPF programs can integrate more closely with the
-    networking stack's understanding of the traffic transiting the kernel.
+    %To provide access to socket state from a BPF helper call, the verifier
+    %needs the ability to track the safety of the calls, including appropriate
+    %reference counting upon the underlying socket for the duration of the
+    %access.
+
+    %With such an ability ability in the verifier, we add socket lookup and
+    %release function calls to the BPF API, allowing BPF programs to safely find
+    %a socket and build logic upon the presence or attributes of a socket.
+    %Applications for this functionality range from load-balancing traffic based
+    %on the presence of a listening application to building stateful firewalling
+    %primitives to understand whether traffic for this connection has been seen
+    %before. With this new functionality, BPF programs can integrate more
+    %closely with the networking stack's understanding of the traffic transiting
+    %the kernel.
+
+    %provide safe access to sockets from BPF packet handling hooks. The BPF verifier
+    %is extended to gain understanding of reference-counted pointer types, which
+    %allows BPF code paths to
 
@@
 
-\begin{itemize}
-\item BPF started with sockets, but only provided access to packets
-\item Previous helper extensions allowed e.g. socket operation handling
-\item Arbitrary socket lookup from socket/packet hooks allows more powerful BPF
-      programs
-\item Use cases revolve around building logic in socket/packet layer hooks that
-      change the behaviour based on the existence and attributes of a socket.
-\end{itemize}
+% Over the past several years, BPF has steadily become more powerful in
+% multiple ways: Through building more intelligence into the verifier which
+% allows more complex programs to be loaded, and through extension of the API
+% such as by adding new map types and new native BPF function calls. While
+% BPF has its roots in applying filters at the socket layer, the ability to
+% introspect the sockets relating to traffic being filtered has been limited.
+
+To build such awareness into a BPF helper, the verifier needs the ability to
+track the safety of the calls, including appropriate reference counting upon
+the underlying socket. This talk walks through extensions to the verifier to
+perform tracking of references in a BPF program. This allows BPF developers to
+extend the UAPI with functions that allocate and release resources within the
+execution lifetime of a BPF program, and the verifier will validate that the
+resources are released exactly once prior to program completion.
+
+Using this new reference tracking ability in the verifier, we add socket lookup
+and release function calls to the BPF API, allowing BPF programs to safely find
+a socket and build logic upon the presence or attributes of a socket. This can
+be used to load-balance traffic based on the presence of a listening
+application, or to implement stateful firewalling primitives to understand
+whether traffic for this connection has been seen before. With this new
+functionality, BPF programs can integrate more closely with the networking
+stack's understanding of the traffic transiting the kernel.
+
+\todo{Desire to target logic based upon socket existence/attributes at packet
+layers}
 
@@
 that would be incompatible with the core socket handling logic. This is
-provided through extension of existing functionality which performs bounds
+provided through extension of existing functionality which provides bounds
 verification and offset conversion for programs operating on CGroups hooks.
-Additionally, the underlying memory must remain associated with the socket for
-the duration of accesses from the BPF program, which may be guaranteed by
-taking a reference on sockets for the duration of their use. The following
-sections explain how this is implemented through the introduction of a new
-pointer type in the verifier, and through the extension of the verifier to
-recognise assembly instructions that represent reference acquisition and
-release.
+Additionally, the underlying socket memory must remain associated with the
+socket for the duration of accesses from the BPF program, which may be
+guaranteed by taking a reference on sockets for the duration of their use.
 
@@
 
-Sockets in Linux are reference-counted to track usage and handle memory
-release. In general to ensure the safety of socket for the duration of access
-in BPF programs, this reference count needs to be incremented when sockets are
-referenced from BPF programs. Two options were considered for how to guarantee
-this: Implicit reference tracking and explicit reference tracking.
+As with many other structure types, Linux sockets are reference-counted to
+track usage and handle memory release. In general, to ensure the safety of
+socket for the duration of access in BPF programs, when sockets are referenced
+from BPF programs the reference count for the socket needs to be incremented.
+Two options were considered for how to guarantee this: Implicit reference
+tracking and explicit reference tracking.
 
-\subsubsection{Implicit reference tracking.}
+Implicit reference tracking calls for the BPF infrastructure to handle
+reference tracking, and to hide this detail from the BPF API. Whenever a socket
+lookup function is called, a reference is taken and the socket is added to a
+reference list. At the end of the BPF program execution, the core code could
+walk this reference list and release each reference. However, collecting
+references and releasing them at the end of the BPF program invocation has the
+unfortunate overhead that even the execution of BPF programs that do not make
+use of the new socket lookup helpers would need to pay the cost of checking the
+list of references to free. When dealing with sufficiently demanding use cases
+such as those which use XDP, even the few instructions required to implement
+such a check may have a measurable impact on performance.
 
-This calls for the BPF infrastructure to handle reference tracking, and to hide
-this detail from the BPF API. Whenever a socket lookup function is called, a
-reference is taken and the socket is added to a reference list. At the end of
-the BPF program execution, the core code could walk this reference list and
-release each reference. However, collecting references and releasing them at
-the end of the BPF program invocation has the unfortunate overhead that even
-the execution of BPF programs that do not make use of the new socket lookup
-helpers would need to pay the cost of checking the list of references to free.
-When dealing with sufficiently demanding use cases such as those which use XDP,
-even the few instructions required to implement such a check may have a
-measurable impact on performance.
+Explicit reference tracking calls for reference acquisition and release
+semantics to be built into the API. If references must be taken, then requiring
+BPF program writers to explicitly handle these references ensures that the
+program writer must understand the potential impact this may have on the
+operation of the program (including atomic operation cost), and it also ensures
+that the cost is only borne by programs that make use of this feature.
 
-\subsubsection{Explicit reference tracking.}
-
-This calls for reference acquisition and release semantics to be built into the
-API. If references must be taken, then requiring BPF program writers to
-explicitly handle these references ensures that the program writer must
-understand the potential impact this may have on the operation of the program
-(including atomic operation cost), and it also ensures that the cost is only
-borne by programs that make use of this feature.
-
-\subsubsection{Implementation.}
-
-Based upon the tradeoffs described above, explicit reference tracking was
-chosen. From an implementation perspective, this requires tracking pointer
+\todo{Lead into the implementation} This safety is provided by tracking pointer
 values through each conditional path in the program, and rejecting the load of
 programs that fail to "balance" resource acquisition (lookup) with release.
-This implementation works as follows.
 
@@
 in the verifier to associate the acquisition and release functions. When the
-instruction that acquires a resource is processed, a resource identifier is
-allocated for this resource. This identifier is kept in a list in the verifier
-state, and it is also associated with the register which receives the resulting
-resource pointer. When processing a corresponding release helper call, if a
-parameter to the function contains a register that is associated with a
-resource identifier, then the identifier is removed from the verifier state. If
-any paths in the program reach the final (\verb+BPF_EXIT+) instruction and the
-verifier state contains any resource identifier association, then the program
-is considered unsafe as it has leaked the resource.
-
-Some instruction calls are restricted while holding a resource reference to
-avoid leaking references, for example the \verb+bpf_tail_call+ helper function.
+instruction that acquires a resource is processed, the verifier state is
+updated to include a resource identifier for the resulting value, and the
+register's value is associated with this identifier. When processing a
+corresponding release helper call, if a parameter to the function contains a
+resource identifier, then the corresponding identifier is removed from the
+verifier state. If any paths in the program reach the final (\verb+BPF_EXIT+)
+instruction and the verifier state still contains resource identifiers, then
+the program is considered unsafe as it has leaked the resource.
 
@@
 
-The following considerations were made for implementing the lookup helpers. The
-consideration of each of these items reflects the definition of the lookup
-helpers in Listing~\ref{lst:helper-api}.
+The following considerations were made for implementing the lookup helpers:
 
@@
 \end{itemize}
+
+The consideration of each of these items reflects the definition of the lookup
+helpers (Listing~\ref{lst:helper-api}):
 
@@
 
-Multiple optimizations were proposed during development of this feature to
-reduce the runtime overhead of using the socket lookup helpers. Two such
-optimizations are described below: Skipping reference counting when unnecessary
-to save on atomic instructions, and allowing the use of direct packet pointers
-for the tuple.
-
 \subsubsection{Avoid reference counting when unnecessary.}
@@
 
-The implementation introduced in the 4.20 release cycle includes support for
-looking up TCP and UDP sockets for IPv4 and IPv6 traffic, using the
-\verb+SCHED_CLS+, \verb+SCHED_ACT+ and \verb+SK_SKB+ hook points. A proposed
-patchset extends this to allow the same helpers to be used from the \verb+XDP+
-hook points~\cite{sk-lookup-xdp}.
-
-\todo{Associate custom metadata with socket?}
+The implementation intruduced in the 4.20 release cycle includes support for
+looking up TCP and UDP sockets for IPv4 and IPv6 traffic, using the \verb+tc+
+and \verb+sk_skb+ hook points. A proposed patchset extends this to allow the
+same helpers to be used from the \verb+xdp+ hook points.
 
@@
 
-Stateful firewalling is distinguished from stateless firewalling in that it
-provides the ability to associate two directions of a connection together and
-apply a filtering policy based upon the direction of the connection. This
-commonly involves using a connection tracker to reconstruct the connection
-state of each endpoint of the connection without specific knowledge from those
-endpoints~\cite{ovs-ct,conntrack}. In cases where the Linux stack acts as a
-firewall on a node in the network that is separate from both endpoints, In many
-cases this model is used when filtering traffic where one endpoint is colocated
-with the Linux stack that performs the firewalling, which already contains
-information about the local end of the connection in a slightly.
-
 \subsection{Load-balancing}

### 2018-10-27 paper/paper.tex
paper: Switch back to original abstract

--- 2018-10-27 paper/paper.tex
@@
 
-Sockets in Linux are reference-counted to track usage and handle memory
-release. In general to ensure the safety of socket for the duration of access
-in BPF programs, this reference count needs to be incremented when sockets are
-referenced from BPF programs. Two options were considered for how to guarantee
-this: Implicit reference tracking and explicit reference tracking.
+As with many other structure types, Linux sockets are reference-counted to
+track usage and handle memory release. In general to ensure the safety of
+socket for the duration of access in BPF programs, this reference count needs
+to be incremented when sockets are referenced from BPF programs. Two options
+were considered for how to guarantee this: Implicit reference tracking and
+explicit reference tracking.
 

### 2018-10-27 paper/paper.tex
Add \todo command

--- 2018-10-27 paper/paper.tex
@@
 \pdfinfo{
-/Title (Title)
-/Author (Author)}
-\title{Title}
-\author{Author\\
-Affiliation\\
-Email\\
-}
+/Title (Building socket-aware BPF programs)
+/Author (Joe Stringer)}
+
+\title{Building socket-aware BPF programs}
+\author{Joe Stringer}
+\affiliation{%
+      \institution{Cilium.io}}
+\email{joe@cilium.io}
 
@@
 
+\usepackage{xcolor}
+\newcommand\todo[1]{\textcolor{red}{#1}}
+
 \begin{document}
+
+\begin{abstract}
+
+Over the past several years, BPF has steadily become more powerful in
+multiple ways: Through building more intelligence into the verifier which
+allows more complex programs to be loaded, and through extension of the API
+such as by adding new map types and new native BPF function calls. While
+BPF has its roots in applying filters at the socket layer, the ability to
+introspect the sockets relating to traffic being filtered has been limited.
+
+To build such awareness into a BPF helper, the verifier needs the ability to
+track the safety of the calls, including appropriate reference counting upon
+the underlying socket. This talk walks through extensions to the verifier to
+perform tracking of references in a BPF program. This allows BPF developers to
+extend the UAPI with functions that allocate and release resources within the
+execution lifetime of a BPF program, and the verifier will validate that the
+resources are released exactly once prior to program completion.
+
+Using this new reference tracking ability in the verifier, we add socket lookup
+and release function calls to the BPF API, allowing BPF programs to safely find
+a socket and build logic upon the presence or attributes of a socket. This can
+be used to load-balance traffic based on the presence of a listening
+application, or to implement stateful firewalling primitives to understand
+whether traffic for this connection has been seen before. With this new
+functionality, BPF programs can integrate more closely with the networking
+stack's understanding of the traffic transiting the kernel.
+
+\end{abstract}
 
@@
 
-\begin{abstract}
-
-\end{abstract}
-
 \section{Keywords}
 
+BPF, Linux, packet processing, sockets
+
 \section{Introduction}
-
-\section{Conclusion}
 
@@
 \bibliographystyle{plainnat}
-%\bibliography{paper}
+\bibliography{paper}
 

### 2018-10-27 paper/paper.tex
paper: Fix up API listing

--- 2018-10-27 paper/paper.tex
@@
 \newcommand\todo[1]{\textcolor{red}{#1}}
+\begin{filecontents*}{sk-lookup-api.c}
+\import{sk-lookup-api.c}
+\end{filecontents*}
 
@@
 
-\todo{Span this listing across two columns}
-\begin{lstlisting}[caption={BPF API helper functions for socket lookup},language=c,label=lst:helper-api,breaklines=true]
-struct bpf_sock *bpf_sk_lookup_tcp(void *ctx, struct bpf_sock_tuple *tuple, u32 tuple_size,
-                                   u32 netns, u64 flags)
-struct bpf_sock *bpf_sk_lookup_udp(void *ctx, struct bpf_sock_tuple *tuple, u32 tuple_size,
-                                   u32 netns, u64 flags)
-int bpf_sk_release(struct bpf_sock *sk)
-\end{lstlisting}
+\lstinputlisting[float=*t,%
+  caption={BPF API helper functions for socket lookup},%
+  language=c,%
+  label=lst:helper-api,%
+  breaklines=true]{sk-lookup-api.c}
 

### 2018-10-27 sk-lookup-api.c

@@ -1,0 +1,5 @@
+struct bpf_sock *bpf_sk_lookup_tcp(void *ctx, struct bpf_sock_tuple *tuple,
+                                   u32 tuple_size, u32 netns, u64 flags);
+struct bpf_sock *bpf_sk_lookup_udp(void *ctx, struct bpf_sock_tuple *tuple,
+                                   u32 tuple_size, u32 netns, u64 flags);
+void bpf_sk_release(struct bpf_sock *sk);

### 2018-10-27 paper/paper.tex
paper: Describe socket pointer awareness

--- 2018-10-27 paper/paper.tex
@@
 
+    This paper presents the changes necessary to extend socket introspection
+    capabilities of the Linux kernel's BPF packet hooks. To provide safety
+    guarantees around the access of the socket, the verifier was extended to
+    track allocation and release of kernel resources. Using this mechanism, we
+    introduce socket lookup and release helpers into the BPF API.
+
+    %To provide access to socket state from a BPF helper call, the verifier
+    %needs the ability to track the safety of the calls, including appropriate
+    %reference counting upon the underlying socket for the duration of the
+    %access.
+
+    %With such an ability ability in the verifier, we add socket lookup and
+    %release function calls to the BPF API, allowing BPF programs to safely find
+    %a socket and build logic upon the presence or attributes of a socket.
+    %Applications for this functionality range from load-balancing traffic based
+    %on the presence of a listening application to building stateful firewalling
+    %primitives to understand whether traffic for this connection has been seen
+    %before. With this new functionality, BPF programs can integrate more
+    %closely with the networking stack's understanding of the traffic transiting
+    %the kernel.
+
+    %provide safe access to sockets from BPF packet handling hooks. The BPF verifier
+    %is extended to gain understanding of reference-counted pointer types, which
+    %allows BPF code paths to
+
+\end{abstract}
+
+\maketitle
+
+\section{Keywords}
+
+BPF, Linux, packet processing, sockets
+
+\section{Introduction}
+
+% Over the past several years, BPF has steadily become more powerful in
+% multiple ways: Through building more intelligence into the verifier which
+% allows more complex programs to be loaded, and through extension of the API
+% such as by adding new map types and new native BPF function calls. While
+% BPF has its roots in applying filters at the socket layer, the ability to
+% introspect the sockets relating to traffic being filtered has been limited.
+
 To build such awareness into a BPF helper, the verifier needs the ability to
@@
 
-\end{abstract}
+This safety is provided by tracking pointer values through each conditional
+path in the program, and rejecting the load of programs that fail to "balance"
+resource acquisition (lookup) with release.
 
-\maketitle
+\section{Background}
 
-\section{Keywords}
+The BPF verifier must ensure that all BPF programs that are loaded into the
+kernel are safe to execute. For proposed access into socket types, this
+requires validation that memory accesses within the socket memory are
+appropriate with respect to the bounds of the structure and the norms of
+read/write access, as well as guaranteeing that underlying socket memory
+remains associated with the socket for the duration of accesses from the BPF
+program.
 
-BPF, Linux, packet processing, sockets
+For some socket types, such as UDP sockets, or TCP listen sockets, memory
+access safety can be achieved with minimal implementation: The destruction of
+such sockets is already governed by standard RCU rules, meaning that while the
+RCU lock is held, they can be safely referenced; Once the RCU grace period is
+reached, the memory may be freed and references to the socket are no longer
+safe to use. Because BPF programs run under the RCU lock, this allows simple,
+safe access of properties of these sockets without requiring additional
+mechanisms to ensure the safety of the socket pointers.
 
-\section{Introduction}
+When it comes to TCP sockets that are not governed by RCU, the networking stack
+uses reference counting to manage socket memory instead. To ensure that the
+socket remains safe to access for a period of BPF program execution, the BPF
+program must also take a reference on the socket. This also implies that such
+references must be subsequently released to prevent leaking kernel resources.
+
+\subsection{Pointers to Sockets}
+
+Earlier work~\cite{bpf-sock} introduced a \verb+bpf_sock+ structure to the BPF
+API and implemented bounds and access type checking, along with offset
+rewriting for converting the access of socket attributes from BPF instructions
+into equivalent accesses of the underlying kernel types. This work was
+introduced with new BPF hook points which provide the socket structure as the
+context of the BPF program.
+
+To allow BPF programs to retrieve and access a socket pointer, the verifier
+must be made aware of when a register contains a pointer of this type, and also
+how to validate pointer access. The existing verifier logic handles context
+pointers in a generic manner via \verb+bpf_verifier_ops+. However, the packet
+event hook points targeted in this paper already contain a context pointer of a
+type that is different from \verb+bpf_sock+, so some additional logic was
+required in the verifier to understand this pointer type. The patch series
+associated with this paper introduced a new pointer type specific to sockets,
+and linked its validation directly to the aforementioned socket verification
+functions without using the verifier operations abstraction.
+
+With the verifier now aware of socket pointer types, this could be built upon
+to implicitly associate socket pointers with references to kernel resources.
+
+\subsection{Reference tracking}
+
+Alternatives:
+\begin{itemize}
+\item RCU-protect other socket types
+\item Collect references during program execution and release afterwards
+\item Build acquire/release semantics into the API
+\end{itemize}
+
+Collecting references and releasing afterwards has the unfortunate overhead
+that even the execution of BPF programs that do not make use of the new socket
+lookup helpers would need to pay the cost of checking the list of references
+to free. When dealing with sufficiently demanding use cases such as those which
+use XDP, even the few instructions required to implement such a check may have
+a measurable impact on performance.
+
+Calls to helper functions which acquire references to resources are annotated
+in the verifier to associate the acquisition and release functions. When the
+instruction that acquires a resource is processed, the verifier state is
+updated to include a resource identifier for the resulting value, and the
+register's value is associated with this identifier. When processing a
+corresponding release helper call, if a parameter to the function contains a
+resource identifier, then the corresponding identifire is removed from the
+verifier state. If any paths in the program reach the final (\verb+BPF_EXIT+)
+instruction and the verifier state still contains resource identifiers, then
+the program is considered unsafe as it has leaked the resource.
+
+\subsection{Runtime protection}
+
+\todo{LD\_ABS, LD\_IND}
+
+\section{Extending the BPF API}
+
+\section{Use cases}
+
+\subsection{Stateful Firewalling}
+
+\subsection{Load-balancing}
+
+\section{Conclusion}
+
+This paper describes the contribution of two extensions to the Linux kernel BPF
+implementation, firstly the introduction of reference tracking logic into the
+verifier, which provides useful base infrastructure for supporting acquire and
+release semantics for kernel resources, and secondly the socket lookup API
+which may be now used from packet paths to allow BPF writers to build logic
+based upon introspection of kernel sockets.
 
 \section{Acknowledgments}
+
+The core BPF maintainers for their guidance and review of the patches.
 

### 2018-10-27 paper/paper.tex
paper: Minor wording suggestions

--- 2018-10-27 paper/paper.tex
@@
 \newcommand\todo[1]{\textcolor{red}{#1}}
-\begin{filecontents*}{sk-lookup-api.c}
-\import{sk-lookup-api.c}
-\end{filecontents*}
 
@@
 
-    To build such awareness into a BPF helper, the verifier needs the ability
-    to track the safety of the calls, including appropriate reference counting
-    upon the underlying socket. This paper describes extensions to the verifier
-    to perform tracking of references in a BPF program. This allows BPF
-    developers to extend the UAPI with functions that allocate and release
-    resources within the execution lifetime of a BPF program, and the verifier
-    will validate that the resources are released exactly once prior to program
-    completion.
+    This paper presents the changes necessary to extend socket introspection
+    capabilities of the Linux kernel's BPF packet hooks. To provide safety
+    guarantees around the access of the socket, the verifier is extended to
+    track allocation and release of kernel resources. Using this mechanism, we
+    introduce socket lookup and release helpers into the BPF API.
 
-    Using this new reference tracking ability in the verifier, we add socket
-    lookup and release function calls to the BPF API, allowing BPF programs to
-    safely find a socket and build logic upon the presence or attributes of a
-    socket. This can be used to load-balance traffic based on the presence of a
-    listening application, or to implement stateful firewalling primitives to
-    understand whether traffic for this connection has been seen before. With
-    this new functionality, BPF programs can integrate more closely with the
-    networking stack's understanding of the traffic transiting the kernel.
+    %These new primitives open up new possibilities for various use cases
+    %including stateful firewalling and load balancing.
+
+    %To provide access to socket state from a BPF helper call, the verifier
+    %needs the ability to track the safety of the calls, including appropriate
+    %reference counting upon the underlying socket for the duration of the
+    %access.
+
+    %With such an ability ability in the verifier, we add socket lookup and
+    %release function calls to the BPF API, allowing BPF programs to safely find
+    %a socket and build logic upon the presence or attributes of a socket.
+    %Applications for this functionality range from load-balancing traffic based
+    %on the presence of a listening application to building stateful firewalling
+    %primitives to understand whether traffic for this connection has been seen
+    %before. With this new functionality, BPF programs can integrate more
+    %closely with the networking stack's understanding of the traffic transiting
+    %the kernel.
+
+    %provide safe access to sockets from BPF packet handling hooks. The BPF verifier
+    %is extended to gain understanding of reference-counted pointer types, which
+    %allows BPF code paths to
 
@@
 \item Previous helper extensions allowed e.g. socket operation handling
-\item Arbitrary socket lookup from socket/packet hooks allows more powerful BPF
-      programs
-\item Use cases revolve around building logic in socket/packet layer hooks that
-      change the behaviour based on the existence and attributes of a socket.
+\item Arbitrary socket lookup from socket/packet lookups allows more powerful
+      BPF programs
 \end{itemize}
+
+% Over the past several years, BPF has steadily become more powerful in
+% multiple ways: Through building more intelligence into the verifier which
+% allows more complex programs to be loaded, and through extension of the API
+% such as by adding new map types and new native BPF function calls. While
+% BPF has its roots in applying filters at the socket layer, the ability to
+% introspect the sockets relating to traffic being filtered has been limited.
+
+To build such awareness into a BPF helper, the verifier needs the ability to
+track the safety of the calls, including appropriate access into socket pointer
+types, and reference counting upon the underlying socket. This paper walks
+through extensions to the verifier to perform tracking of references in a BPF
+program. This allows BPF developers to extend the API with functions that
+allocate and release resources within the execution lifetime of a BPF program,
+and the verifier will validate that the resource acquisition and release is
+balanced across all execution paths of the BPF programs.
+
+Using this new reference tracking ability in the verifier, we add socket lookup
+and release function calls to the BPF API, allowing BPF programs to safely find
+a socket and build logic upon the presence or attributes of a socket. This can
+be used to load-balance traffic based on the presence of a listening
+application, or to implement stateful firewalling primitives to understand
+whether traffic for this connection has been seen before. With this new
+functionality, BPF programs can integrate more closely with the networking
+stack's understanding of the traffic transiting the kernel.
+
+\todo{Desire to target logic based upon socket existence/attributes at packet
+layers}
 
@@
 
-\lstinputlisting[float=*t,%
-  caption={BPF API helper functions for socket lookup},%
-  language=c,%
-  label=lst:helper-api,%
-  breaklines=true]{sk-lookup-api.c}
+The consideration of each of these items reflects the definition of the lookup
+helpers (Listing~\ref{lst:helper-api}):
+
+\todo{Span this listing across two columns}
+\begin{lstlisting}[caption={BPF API helper functions for socket lookup},language=c,label=lst:helper-api,breaklines=true]
+struct bpf_sock *bpf_sk_lookup_tcp(void *ctx, struct bpf_sock_tuple *tuple, u32 tuple_size,
+                                   u32 netns, u64 flags)
+struct bpf_sock *bpf_sk_lookup_udp(void *ctx, struct bpf_sock_tuple *tuple, u32 tuple_size,
+                                   u32 netns, u64 flags)
+int bpf_sk_release(struct bpf_sock *sk)
+\end{lstlisting}
 

### 2018-10-27 paper/paper.tex
paper: Tidy up wording

--- 2018-10-27 paper/paper.tex
@@
 
-\lstinputlisting[float=*t,%
-  caption={BPF API helper functions for socket lookup},%
-  language=c,%
-  label=lst:helper-api,%
-  breaklines=true]{sk-lookup-api.c}
+\todo{Span this listing across two columns}
+\begin{lstlisting}[caption={BPF API helper functions for socket lookup},language=c,label=lst:helper-api,breaklines=true]
+struct bpf_sock *bpf_sk_lookup_tcp(void *ctx, struct bpf_sock_tuple *tuple, u32 tuple_size,
+                                   u32 netns, u64 flags)
+struct bpf_sock *bpf_sk_lookup_udp(void *ctx, struct bpf_sock_tuple *tuple, u32 tuple_size,
+                                   u32 netns, u64 flags)
+int bpf_sk_release(struct bpf_sock *sk)
+\end{lstlisting}
 

### 2018-10-27 paper/paper.tex
paper: Improve general structure and wording

--- 2018-10-27 paper/paper.tex
@@
 As with many other structure types, Linux sockets are reference-counted to
-track usage and handle memory release. In general, to ensure the safety of
-socket for the duration of access in BPF programs, when sockets are referenced
-from BPF programs the reference count for the socket needs to be incremented.
-Two options were considered for how to guarantee this: Implicit reference
-tracking and explicit reference tracking.
+track usage and handle memory release. In general to ensure the safety of
+socket for the duration of access in BPF programs, this reference count needs
+to be incremented when sockets are referenced from BPF programs. Two options
+were considered for how to guarantee this: Implicit reference tracking and
+explicit reference tracking.
 
-Implicit reference tracking calls for the BPF infrastructure to handle
-reference tracking, and to hide this detail from the BPF API. Whenever a socket
-lookup function is called, a reference is taken and the socket is added to a
-reference list. At the end of the BPF program execution, the core code could
-walk this reference list and release each reference. However, collecting
-references and releasing them at the end of the BPF program invocation has the
-unfortunate overhead that even the execution of BPF programs that do not make
-use of the new socket lookup helpers would need to pay the cost of checking the
-list of references to free. When dealing with sufficiently demanding use cases
-such as those which use XDP, even the few instructions required to implement
-such a check may have a measurable impact on performance.
+\subsubsection{Implicit reference tracking.}
 
-Explicit reference tracking calls for reference acquisition and release
-semantics to be built into the API. If references must be taken, then requiring
-BPF program writers to explicitly handle these references ensures that the
-program writer must understand the potential impact this may have on the
-operation of the program (including atomic operation cost), and it also ensures
-that the cost is only borne by programs that make use of this feature.
+This calls for the BPF infrastructure to handle reference tracking, and to hide
+this detail from the BPF API. Whenever a socket lookup function is called, a
+reference is taken and the socket is added to a reference list. At the end of
+the BPF program execution, the core code could walk this reference list and
+release each reference. However, collecting references and releasing them at
+the end of the BPF program invocation has the unfortunate overhead that even
+the execution of BPF programs that do not make use of the new socket lookup
+helpers would need to pay the cost of checking the list of references to free.
+When dealing with sufficiently demanding use cases such as those which use XDP,
+even the few instructions required to implement such a check may have a
+measurable impact on performance.
 
-\todo{Lead into the implementation} This safety is provided by tracking pointer
+\subsubsection{Explicit reference tracking.}
+
+This calls for reference acquisition and release semantics to be built into the
+API. If references must be taken, then requiring BPF program writers to
+explicitly handle these references ensures that the program writer must
+understand the potential impact this may have on the operation of the program
+(including atomic operation cost), and it also ensures that the cost is only
+borne by programs that make use of this feature.
+
+\subsubsection{Implementation.}
+
+Based upon the tradeoffs described above, explicit reference tracking was
+chosen. From an implementation perspective, this requires tracking pointer
 values through each conditional path in the program, and rejecting the load of
 programs that fail to "balance" resource acquisition (lookup) with release.
+This implementation works as follows.
 
@@
 in the verifier to associate the acquisition and release functions. When the
-instruction that acquires a resource is processed, the verifier state is
-updated to include a resource identifier for the resulting value, and the
-register's value is associated with this identifier. When processing a
-corresponding release helper call, if a parameter to the function contains a
-resource identifier, then the corresponding identifier is removed from the
-verifier state. If any paths in the program reach the final (\verb+BPF_EXIT+)
-instruction and the verifier state still contains resource identifiers, then
-the program is considered unsafe as it has leaked the resource.
+instruction that acquires a resource is processed, a resource identifier is
+allocated for this resource. This identifier is kept in a list in the verifier
+state, and it is also associated with the register which receives the resulting
+resource pointer. When processing a corresponding release helper call, if a
+parameter to the function contains a register that is associated with a
+resource identifier, then the identifier is removed from the verifier state. If
+any paths in the program reach the final (\verb+BPF_EXIT+) instruction and the
+verifier state contains any resource identifier association, then the program
+is considered unsafe as it has leaked the resource.
+
+Some instruction calls are restricted while holding a resource reference to
+avoid leaking references, for example the \verb+bpf_tail_call+ helper function.
 
@@
 
-The following considerations were made for implementing the lookup helpers:
+The following considerations were made for implementing the lookup helpers. The
+consideration of each of these items reflects the definition of the lookup
+helpers in Listing~\ref{lst:helper-api}.
 
@@
 \end{itemize}
-
-The consideration of each of these items reflects the definition of the lookup
-helpers (Listing~\ref{lst:helper-api}):
 
@@
 
+Multiple optimizations were proposed during development of this feature to
+reduce the runtime overhead of using the socket lookup helpers. Two such
+optimizations are described below: Skipping reference counting when unnecessary
+to save on atomic instructions, and allowing the use of direct packet pointers
+for the tuple.
+
 \subsubsection{Avoid reference counting when unnecessary.}
@@
 The implementation intruduced in the 4.20 release cycle includes support for
-looking up TCP and UDP sockets for IPv4 and IPv6 traffic, using the \verb+tc+
-and \verb+sk_skb+ hook points. A proposed patchset extends this to allow the
-same helpers to be used from the \verb+xdp+ hook points.
+looking up TCP and UDP sockets for IPv4 and IPv6 traffic, using the
+\verb+SCHED_CLS+, \verb+SCHED_ACT+ and \verb+SK_SKB+ hook points. A proposed
+patchset extends this to allow the same helpers to be used from the \verb+XDP+
+hook points~\cite{sk-lookup-xdp}.
 

### 2018-10-27 paper/paper.tex
paper: Tidy up outro

@@
 
-The implementation intruduced in the 4.20 release cycle includes support for
+The implementation introduced in the 4.20 release cycle includes support for
 looking up TCP and UDP sockets for IPv4 and IPv6 traffic, using the
@@
 hook points~\cite{sk-lookup-xdp}.
+
+\todo{Associate custom metadata with socket?}
 
@@
 
-This paper describes the contribution of two extensions to the Linux kernel BPF
-implementation, firstly the introduction of reference tracking logic into the
-verifier, which provides useful base infrastructure for supporting acquire and
-release semantics for kernel resources, and secondly the socket lookup API
-which may be now used from packet paths to allow BPF writers to build logic
-based upon introspection of kernel sockets.
+This paper describes the contribution of new BPF verifier functionality in the
+Linux kernel to track references to kernel resources and uses these to provide
+access to socket introspection capabilities from packet handling hooks. The
+introduction of reference tracking logic into the verifier provides
+useful base infrastructure for supporting acquire and release semantics for
+kernel resources which may prove useful for other helpers in future, and the
+socket lookup API makes the packet hooks more powerful to support use cases
+such as stateful firewalling or load-balancing based on the sockets that are
+open on a system.
 
@@
 
-The core BPF maintainers for their guidance and review of the patches.
+The core BPF maintainers provided great guidance and review of the patches, in
+particular Alexei Starovoitov, Daniel Borkmann and Martin KaFai Lau.
 

### 2018-10-27 paper/paper.tex
paper: Add structures listings

--- 2018-10-27 paper/paper.tex
@@
 \newcommand\todo[1]{\textcolor{red}{#1}}
+
+\begin{filecontents*}{sk-lookup-api.c}
+\import{sk-lookup-api.c}
+\end{filecontents*}
+\begin{filecontents*}{sk-lookup-structures.c}
+\import{sk-lookup-structures.c}
+\end{filecontents*}
 
@@
 consideration of each of these items reflects the definition of the lookup
-helpers in Listing~\ref{lst:helper-api}.
+helpers in Listing~\ref{lst:helper-api} and the structure definitions in
+Listing~\ref{lst:helper-struct}.
 
@@
 
-\todo{Span this listing across two columns}
-\begin{lstlisting}[caption={BPF API helper functions for socket lookup},language=c,label=lst:helper-api,breaklines=true]
-struct bpf_sock *bpf_sk_lookup_tcp(void *ctx, struct bpf_sock_tuple *tuple, u32 tuple_size,
-                                   u32 netns, u64 flags)
-struct bpf_sock *bpf_sk_lookup_udp(void *ctx, struct bpf_sock_tuple *tuple, u32 tuple_size,
-                                   u32 netns, u64 flags)
-int bpf_sk_release(struct bpf_sock *sk)
-\end{lstlisting}
+\lstinputlisting[float=*b,%
+  caption={BPF API helper functions for socket lookup},%
+  language=c,%
+  label=lst:helper-api,%
+  breaklines=true]{sk-lookup-api.c}
+
+\lstinputlisting[float=*b,%
+  caption={BPF API structures for socket lookup (Linux v4.20)},%
+  language=c,%
+  label=lst:helper-struct,%
+  breaklines=true]{sk-lookup-structures.c}
 

### 2018-10-27 paper/sk-lookup-structures.c

@@ -1,0 +1,34 @@
+struct bpf_sock_tuple {
+	union {
+		struct {
+			__be32 saddr;
+			__be32 daddr;
+			__be16 sport;
+			__be16 dport;
+		} ipv4;
+		struct {
+			__be32 saddr[4];
+			__be32 daddr[4];
+			__be16 sport;
+			__be16 dport;
+		} ipv6;
+	};
+};
+
+struct bpf_sock {
+	__u32 bound_dev_if;
+	__u32 family;
+	__u32 type;
+	__u32 protocol;
+	__u32 mark;
+	__u32 priority;
+	__u32 src_ip4;		/* Allows 1,2,4-byte read.
+				 * Stored in network byte order.
+				 */
+	__u32 src_ip6[4];	/* Allows 1,2,4-byte read.
+				 * Stored in network byte order.
+				 */
+	__u32 src_port;		/* Allows 4-byte read.
+				 * Stored in host byte order
+				 */
+};

### 2018-10-27 paper/paper.tex
paper: More fleshing out

@@
 
+%To ensure that the socket remains safe to access for a period of BPF program
+%execution, the BPF program must also take a reference on the socket. This also
+%implies that such references must be subsequently released to prevent leaking
+%kernel resources.
+
 \subsubsection{Implicit reference tracking.}
@@
 helpers would need to pay the cost of checking the list of references to free.
-When dealing with sufficiently demanding use cases such as those which use XDP,
-even the few instructions required to implement such a check may have a
-measurable impact on performance.
+When dealing with sufficiently demanding use cases such as those which use
+eXpress Data Path (XDP)~\cite{xdp}, even the few instructions required to
+implement such a check may have a measurable impact on performance.
 
@@
 Some instruction calls are restricted while holding a resource reference to
-avoid leaking references, for example the \verb+bpf_tail_call+ helper function.
+avoid leaking references, for example the \verb+bpf_tail_call+ helper function,
+which would otherwise leak the pointer reference to a subsequent BPF program.
 
@@
 the program. Therefore, after execution of a BPF program containing resource
-acquire and release functions which are verified by the logic above, the
-underlying resources should not leak.
+acquire and release functions which are verified by the logic above, references
+to those resources should not leak.
 
@@
 prior to the final state. Without mitigation, this could lead to leaking of
-references. Newer BPF APIs provide better alternatives for packet access, so we
-disallow the use of these instructions while holding a socket reference.
+references at runtime, even if the instructions appear to balance acquisition
+and release at verification time. Newer BPF APIs provide better alternatives to
+these instructions for packet access, so we disallow the use of these
+instructions while holding a socket reference.
 
@@
 
-To implement explicit reference tracking, the API required at least two
-functions to handle referencing of sockets from the BPF API: A socket lookup
-function and a socket release function. This section describes the API
-considerations for each of these.
+To provide an API that implements explicit reference tracking, at least two
+functions are needed to handle referencing of sockets: A socket lookup function
+and a socket release function. This section describes the API considerations
+for each of these.
 
@@
 Linux network devices are associated with a particular network namespace which
-provides logical network stack separation. Sockets are also inherently
-associated with an application that operates in a particular network namespace.
-The simplest path for handling namespacing from the lookup handlers would be to
-allow BPF programs operating on a device within a network namespace to only
-find sockets in the same network namespace. By passing the BPF program context
-into the socket lookup helper, the implementation can derive the source network
-namespace from this context. Beyond this, with the growing popularity of
-containers it may be desirable for a network orchestration tool to make use of
-socket properties inside container network namespaces to influence decisions
+provides logical network stack separation~\cite{netns, netns-netdevice}.
+Sockets are also inherently associated with an application that operates in a
+particular network namespace~\cite{netns-sock}. The simplest path for handling
+namespacing from socket lookup handlers operating from a network device would
+be to allow BPF programs operating on a device within a network namespace to
+only find sockets in the same network namespace. By passing the BPF program
+context into the socket lookup helper, the implementation can derive the source
+network namespace from this context. Beyond this, with the growing popularity
+of containers it may be desirable for a network orchestration tool to make use
+of socket properties inside container network namespaces to influence decisions
 made on packets outside the container---for instance, on packets being received
-into a node prior to the packet passing into the container network namespace,
-or if filtering is implemented on a veth pair outside of a container. By
-allowing the netns ID to be specified, this use case can be supported as well.
+into a node prior to passing the packet into the network namespace. By allowing
+the netns ID to be specified, this use case can be supported as well.
 
@@
 metadata. This would however limit the potential uses for such a helper. Two
-cases that would be more limited with this model are usage from XDP, or if the
-BPF program implements a form of network address translation. From XDP, if the
-raw \verb+xdp_md+ context is provided to such a helper, then the helper would
-need to parse the packet to understand the 5-tuple. This would duplicate the
-standard stack paths which are executed after XDP where the \verb+sk_buff+ is
-built, but this information would be subsequently thrown away upon return of
+cases that would be more limited with this model are usage from XDP (which does
+not inherently parse packets or collect packet metadata into the context), or
+if the BPF program implements a form of network address translation. From XDP,
+if the raw \verb+xdp_md+ context is provided to such a helper, then the helper
+would need to parse the packet to understand the 5-tuple. This would duplicate
+the standard stack paths which are executed after XDP where the \verb+sk_buff+
+is built, but this information would be subsequently thrown away upon return of
 the helper call. Adding to this, if the program performs network address
@@
 methods for selecting a socket when the application uses \verb+SO_REUSEPORT+.
-Some discussion on the mailinglist proposed allowing BPF program writers to
-influence the socket selection mechanism~\cite{netdev-sk-select}.
+Some subsequent discussion on the mailinglist proposed allowing BPF program
+writers to influence the socket selection mechanism~\cite{netdev-sk-select}.
 
@@
 TCP or UDP. The initial RFC of this patch series proposed a single lookup
-function which would choose the L4 protocol based upon a field in the
+function which would choose the Layer~4 protocol based upon a field in the
 tuple~\cite{sk-lookup-rfc}, however this would make it more difficult for BPF
 program writers to detect the support for different protocol types at compile
-time. There is not a significant number of supported L4 socket types in the
-Linux stack today, and the number is not expected to increase drastically, so
-it was considered simpler and easier to split out each L4 socket lookup
-function into an independent helper call.
+time. When a socket type is unsupported, the implementation would return
+\verb+NULL+, implying that there is no such socket. There is not a significant
+number of supported Layer~4 socket types in the Linux stack today, and the
+number is not expected to increase drastically, so it was considered simpler
+and easier to split out each Layer~4 socket lookup function into an independent
+helper call, so that the result could simply return a socket or \verb+NULL+ and
+not need to encode a representation of other error conditions.
 
@@
 such sockets is already governed by standard RCU rules, meaning that while the
-RCU lock is held, they can be safely referenced; Once the RCU grace period is
-reached, the memory may be freed and references to the socket are no longer
-safe to use. Because BPF programs run under the RCU lock, this allows simple,
-safe access of properties of these sockets without requiring additional
-mechanisms to ensure the safety of the socket pointers.
+RCU lock is held, they can be safely accessed without holding a reference; Once
+the RCU grace period is reached, the memory may be freed and references to the
+socket are no longer safe to use. For these socket types, since BPF programs
+run under the RCU lock, the properties of these sockets can be accessed
+directly without taking a reference on the socket. As such, the implementation
+of the socket lookup and release functions can avoid the atomic reference count
+increment operation.
 
-When it comes to TCP sockets that are not governed by RCU, the networking stack
-uses reference counting to manage socket memory instead. To ensure that the
-socket remains safe to access for a period of BPF program execution, the BPF
-program must also take a reference on the socket. This also implies that such
-references must be subsequently released to prevent leaking kernel resources.
+The API retains explicit reference tracking to ensure consistency of the API
+and to allow multiple underlying implementations to handle reference tracking
+in a way that is safe for the implementation. For instance, for  TCP sockets
+that are not governed by RCU, the networking stack uses reference counting to
+manage socket memory instead, so the socket lookup and release functions would
+take and release references on the socket, respectively.
 
@@
 immediately followed by the TCP or UDP header without IP options in between,
-then a pointer directly to the L3 addresses may be passed to the lookup
-function, allowing the helper function to directly pull the addresses and ports
-from the packet buffer rather than requiring the BPF program to first extract
-these onto the stack.
+then a pointer to the packet data at the offset of the Layer~3 addresses may be
+passed to the lookup function, allowing the implementation to directly pull the
+addresses and ports from the packet buffer rather than requiring the BPF
+program to first extract these onto the stack and pass a pointer to the stack
+copy of the tuple.
 
@@
 state of each endpoint of the connection without specific knowledge from those
-endpoints~\cite{ovs-ct,conntrack}. In cases where the Linux stack acts as a
-firewall on a node in the network that is separate from both endpoints, In many
-cases this model is used when filtering traffic where one endpoint is colocated
-with the Linux stack that performs the firewalling, which already contains
-information about the local end of the connection in a slightly.
+endpoints~\cite{conntrack,ovs-ct}. In cases where Linux is configured as a
+firewall between applications that exist in distinct Linux kernel copies
+running in the network, this makes sense; The information that is needed to
+perform stateful firewalling is not present on the Linux instance that needs
+this information, so the information must either be shared from the peers
+(assuming that the firewall orchestrator has control over the peer endpoints),
+or this information must be synthesised based upon the packets that are seen on
+this intermediate Linux instance. However, in many cases this model is used
+when filtering traffic where one endpoint is co-located with the Linux stack
+that performs the firewalling, which already contains information about the
+local end of the connection. In this use case, the connection tracking table
+effectively duplicates knowledge that Linux already creates and maintains.
+
+Using the socket lookup functions described in this paper, in conjunction with
+the socket properties including local Layer~3 and Layer~4 information available
+in the \verb+bpf_sock+ (Listing~\ref{lst:helper-struct}), BPF programs can
+identify the directionality of connections for packets ingressing or egressing
+the Linux stack on a network device.
 
@@
 The core BPF maintainers provided great guidance and review of the patches, in
-particular Alexei Starovoitov, Daniel Borkmann and Martin KaFai Lau.
+particular Alexei Starovoitov, Daniel Borkmann and Martin KaFai Lau. Additional
+shout-out to Nitin Hande who provided testing and feedback.
 

### 2018-10-27 paper/paper.tex
paper: Cite flags extensibility LWN article

@@
 New kernel APIs that have any scope for extension should contain a flags
-argument. For the socket lookup helpers, a few ideas had been considered as
-potential future alternatives to the existing behaviour. The lookup functions
-follow the standard socket lookup paths in Linux which have predetermined
-methods for selecting a socket when the application uses \verb+SO_REUSEPORT+.
-Some subsequent discussion on the mailinglist proposed allowing BPF program
-writers to influence the socket selection mechanism~\cite{netdev-sk-select}.
+argument~\cite{syscall-flags}. For the socket lookup helpers, a few ideas had
+been considered as a potential future alternatives to the existing behaviour.
+The lookup functions follow the standard socket lookup paths in Linux which
+have predetermined methods for selecting a socket when the application uses
+\verb+SO_REUSEPORT+. Some subsequent discussion on the mailinglist proposed
+allowing BPF program writers to influence the socket selection
+mechanism~\cite{netdev-sk-select}.
 

### 2018-10-29 paper/paper.tex
Fix and share API headers

@@
 
-\begin{filecontents*}{sk-lookup-api.c}
-\import{sk-lookup-api.c}
+\begin{filecontents*}{../sk-lookup-api.h}
+\import{../sk-lookup-api.h}
 \end{filecontents*}
-\begin{filecontents*}{sk-lookup-structures.c}
-\import{sk-lookup-structures.c}
+\begin{filecontents*}{sk-lookup-structures.h}
+\import{sk-lookup-structures.h}
 \end{filecontents*}
@@
   label=lst:helper-api,%
-  breaklines=true]{sk-lookup-api.c}
+  breaklines=true]{../sk-lookup-api.h}
 
@@
   label=lst:helper-struct,%
-  breaklines=true]{sk-lookup-structures.c}
+  breaklines=true]{sk-lookup-structures.h}
 

### 2018-10-29 paper/sk-lookup-api.c

@@ -1,5 +1,0 @@
-struct bpf_sock *bpf_sk_lookup_tcp(void *ctx, struct bpf_sock_tuple *tuple,
-                                   u32 tuple_size, u32 netns, u64 flags);
-struct bpf_sock *bpf_sk_lookup_udp(void *ctx, struct bpf_sock_tuple *tuple,
-                                   u32 tuple_size, u32 netns, u64 flags);
-void bpf_sk_release(struct bpf_sock *sk);

### 2018-10-29 sk-lookup-api.h

@@ -1,0 +1,8 @@
+struct bpf_sock *
+bpf_sk_lookup_tcp(void *ctx, struct bpf_sock_tuple *tuple,
+                  u32 tuple_size, u32 netns, u64 flags);
+struct bpf_sock *
+bpf_sk_lookup_udp(void *ctx, struct bpf_sock_tuple *tuple,
+                  u32 tuple_size, u32 netns, u64 flags);
+
+void bpf_sk_release(struct bpf_sock *sk);

### 2018-10-29 slides/bpf-sock.h

@@ -1,0 +1,11 @@
+struct bpf_sock {
+        __u32 bound_dev_if;
+        __u32 family;
+        __u32 type;
+        __u32 protocol;
+        __u32 mark;
+        __u32 priority;
+        __u32 src_ip4;           /* NBO */
+        __u32 src_ip6[4];        /* NBO */
+        __u32 src_port;          /* NBO */
+};

### 2018-10-29 slides/bpf-tuple.h

@@ -1,0 +1,16 @@
+struct bpf_sock_tuple {
+        union {
+                struct {
+                        __be32 saddr;
+                        __be32 daddr;
+                        __be16 sport;
+                        __be16 dport;
+                } ipv4;
+                struct {
+                        __be32 saddr[4];
+                        __be32 daddr[4];
+                        __be16 sport;
+                        __be16 dport;
+                } ipv6;
+        };
+};

### 2018-10-29 paper/paper.tex
paper: Turn linux version into a command

@@
 \newcommand\todo[1]{\textcolor{red}{#1}}
+\newcommand\linuxversion{v4.20}
 
@@
 \lstinputlisting[float=*b,%
-  caption={BPF API structures for socket lookup (Linux v4.20)},%
+  caption={BPF API structures for socket lookup (Linux~\linuxversion)},%
   language=c,%

### 2018-10-29 paper/paper.tex
paper: Flesh out introduction

@@
 \newcommand\todo[1]{\textcolor{red}{#1}}
-\newcommand\linuxversion{v4.20}
+\newcommand\linuxversion{{v4.20}}
 
@@
 
-BPF, Linux, packet processing, sockets
+BPF, firewalls, Linux, networks, packet processing, sockets
 
@@
 
-\begin{itemize}
-\item BPF started with sockets, but only provided access to packets
-\item Previous helper extensions allowed e.g. socket operation handling
-\item Arbitrary socket lookup from socket/packet hooks allows more powerful BPF
-      programs
-\item Use cases revolve around building logic in socket/packet layer hooks that
-      change the behaviour based on the existence and attributes of a socket.
-\end{itemize}
+The Berkeley Packet Filter~\cite{cbpf}, introduced in 1992, provided a model
+for the implementation of flexible packet filtering by allowing runtime
+modification of kernel logic from userspace applications. BPF quickly became
+the de facto technology for monitoring the traffic flow within POSIX systems.
+Originally, this was restricted to applying filtering upon sockets attached to
+devices and not to implement filtering capabilities directly for traffic
+flowing through a system. Subsequent work investigated using BPF as a basis for
+packet filtering on devices~\cite{ipfw}, however this design does not appear to
+have been integrated into major operating systems until the introduction of
+extended BPF into Linux~\cite{lwnebpf}.
+
+In recent years, multiple efforts have been made to build new filtering
+implementations based upon
+BPF~\cite{cilium,ovs-ebpf,nsx-harmony,droplet,xdp-ddos,alfwrapper}. Many of
+these implementations aim to provide the functionality of a stateful
+firewall~\cite{stateful-fw} - that is, to allow network policies to be defined
+to allow traffic from one network endpoint to talk to another network endpoint,
+and to implicitly allow responses to pass back through the firewall. Many
+implementations over the past decade have developed this functionality in a
+manner that is agnostic to the locality of the endpoints. This paper revisits
+the possibilities for stateful firewalling when the endpoint socket is
+co-located with the kernel providing firewalling functionality, effectively
+combining the ideas of \cite{ipfw,stateful-fw,dfw}.
+
+This paper is laid out as follows. Firstly, we discuss the constraints that the
+BPF verifier places upon BPF programs and how those constraints translate into
+requirements on the implementation of socket lookup helpers in the Linux BPF
+API. The implementation of reference tracking functionality is presented.
+Secondly, we describe extensions to the BPF API to provide socket introspection
+capabilities, taking into account flexibility and performance. Thirdly, we
+present some use cases for the new functionality, including more detail on the
+stateful firewall case and a description of the use of the new functionality
+for load-balancing. Finally, we conclude and assess the position of this work.
 
@@
 
-The implementation introduced in the 4.20 release cycle includes support for
-looking up TCP and UDP sockets for IPv4 and IPv6 traffic, using the
+The implementation introduced in the Linux~\linuxversion~release cycle includes
+support for looking up TCP and UDP sockets for IPv4 and IPv6 traffic, using the
 \verb+SCHED_CLS+, \verb+SCHED_ACT+ and \verb+SK_SKB+ hook points. A proposed
@@
 
+\todo{Early in the stack, figure out whether the traffic is destined for a
+local process and if so, pass. Otherwise, perform load-balancing to another
+node where the traffic is destined for.}
+
 \section{Conclusion}
@@
 
-The core BPF maintainers provided great guidance and review of the patches, in
-particular Alexei Starovoitov, Daniel Borkmann and Martin KaFai Lau. Additional
-shout-out to Nitin Hande who provided testing and feedback.
+The author would like to acknowledge Alexei Starovoitov, Daniel Borkmann,
+Martin KaFai Lau, Nitin Hande and Thomas Graf for guidance, review and testing
+of the patches.
 

### 2018-10-30 paper/paper.tex
paper: Improve stateful firewalling section
paper: Write up a second use case

@@
 present some use cases for the new functionality, including more detail on the
-stateful firewall case and a description of the use of the new functionality
-for load-balancing. Finally, we conclude and assess the position of this work.
+stateful firewall case and an additional usage in handling management traffic
+in forwarding devices. Finally, we conclude and assess the position of this
+work.
 
@@
 
-\todo{Associate custom metadata with socket?}
+%\todo{Associate custom metadata with socket?}
 
@@
 state of each endpoint of the connection without specific knowledge from those
-endpoints~\cite{conntrack,ovs-ct}. In cases where Linux is configured as a
-firewall between applications that exist in distinct Linux kernel copies
-running in the network, this makes sense; The information that is needed to
-perform stateful firewalling is not present on the Linux instance that needs
-this information, so the information must either be shared from the peers
-(assuming that the firewall orchestrator has control over the peer endpoints),
-or this information must be synthesised based upon the packets that are seen on
-this intermediate Linux instance. However, in many cases this model is used
-when filtering traffic where one endpoint is co-located with the Linux stack
-that performs the firewalling, which already contains information about the
-local end of the connection. In this use case, the connection tracking table
-effectively duplicates knowledge that Linux already creates and maintains.
+endpoints~\cite{conntrack,ovs-ct}.
+
+In cases where Linux is configured as a firewall between applications that
+exist in distinct Linux kernel copies running in the network, this makes sense;
+The information that is needed to perform stateful firewalling is not present
+on the Linux instance that needs this information, so the information must
+either be shared from the peers (assuming that the firewall orchestrator has
+control over the peer endpoints), or this information must be synthesised based
+upon the packets that are seen on this intermediate Linux instance. However, in
+many cases this model is used when filtering traffic where one endpoint is
+co-located with the Linux stack that performs the firewalling, which already
+contains information about the local end of the connection. In this use case,
+the connection tracking table effectively duplicates knowledge that Linux
+already creates and maintains.
 
@@
 identify the directionality of connections for packets ingressing or egressing
-the Linux stack on a network device.
+the Linux stack on a network device. Between this directionality information
+and a network policy provided by a userspace program, BPF programs are able to
+identify locally-sourced or locally-destined traffic and allow or deny traffic
+ingressing or egressing the Linux stack.
 
-\subsection{Load-balancing}
+\subsection{Filtering management traffic in forwarding devices}
 
-\todo{Early in the stack, figure out whether the traffic is destined for a
-local process and if so, pass. Otherwise, perform load-balancing to another
-node where the traffic is destined for.}
+A common design of devices providing routing and forwarding functionality is to
+place high-bandwidth forwarding elements into a system with a low-bandwidth
+CPU. In many cases, the goal of the CPU is to only to handle control traffic
+for configuring the forwarding path, which is a fraction of the traffic handled
+by the actual forwarding path. Existing implementations may direct only traffic
+for a specific IP into the management CPU, or could plausibly use XDP to
+configure DoS filters so that only legitimate traffic is forwarded up the
+stack. Such implementations may not provide the degree of restriction on
+locally destined traffic that is desired, or may require additional
+co-ordination with the socket layer that requires significant orchestration.
+
+Using the upcoming XDP support for socket lookup, early in the driver level of
+the Linux stack the traffic could be compared with the socket table and only
+passed up the stack if the traffic corresponds to a local application. As this
+functionality uses the socket table directly, it is always kept in sync with
+the current applications without additional orchestration of BPF maps. Traffic
+which is not intended to be handled locally can then be either handled by
+additional BPF logic programmed in the XDP hook.
 
@@
 
+%\todo{Associate custom metadata with socket?}
+
 \section{Acknowledgments}

### 2018-10-31 paper/paper.tex
paper: Improve verifier intro

@@
 The BPF verifier must ensure that all BPF programs that are loaded into the
-kernel are safe to execute. For accessing socket pointers, this requires
-validation that memory accesses are within the bounds of the memory allocated
-for the structure, and will not make any modifications to the socket fields
-that would be incompatible with the core socket handling logic. This is
-provided through extension of existing functionality which performs bounds
-verification and offset conversion for programs operating on CGroups hooks.
-Additionally, the underlying memory must remain associated with the socket for
-the duration of accesses from the BPF program, which may be guaranteed by
-taking a reference on sockets for the duration of their use. The following
-sections explain how this is implemented through the introduction of a new
-pointer type in the verifier, and through the extension of the verifier to
-recognise assembly instructions that represent reference acquisition and
-release.
+kernel are safe to execute. This necessitates verification of every path
+through the program to ensure constraint such as that the program will halt,
+and that any dereferencing of pointers is done in a safe and controlled manner.
+For accessing socket pointers, this requires validation that memory accesses
+are within the bounds of the memory allocated for the structure, and that no
+modifications are made to the socket fields that would be cause the core socket
+handling logic to get into a bad state. This is provided through extension of
+existing functionality which performs bounds verification and offset conversion
+for programs operating on CGroups hooks. Additionally, the underlying memory
+must remain associated with the socket for the duration of accesses from the
+BPF program, which may be guaranteed by taking a reference on sockets for the
+duration of their use. The following sections explain how this is implemented
+through the introduction of a new pointer type in the verifier, and through the
+extension of the verifier to recognise assembly instructions that represent
+reference acquisition and release.
 

### 2018-11-01 paper/paper.tex
paper: General flow tidy up

@@
     safely find a socket and build logic upon the presence or attributes of a
-    socket. This can be used to load-balance traffic based on the presence of a
-    listening application, or to implement stateful firewalling primitives to
-    understand whether traffic for this connection has been seen before. With
-    this new functionality, BPF programs can integrate more closely with the
-    networking stack's understanding of the traffic transiting the kernel.
+    socket. This can be used to conditionally redirect traffic based on the
+    presence of a listening application, or to implement stateful firewalling
+    primitives to understand whether traffic for this connection has been seen
+    before. With this new functionality, BPF programs can integrate more
+    closely with the networking stack's understanding of the traffic transiting
+    the kernel.
 
@@
 how to validate pointer access. The existing verifier logic handles context
-pointers in a generic manner via \verb+bpf_verifier_ops+. However, the packet
-event hook points targeted in this paper already contain a context pointer of a
-type that is different from \verb+bpf_sock+, so some additional logic was
-required in the verifier to understand this pointer type. The patch series
-associated with this paper introduced a new pointer type specific to sockets,
-and linked its validation directly to the aforementioned socket verification
-functions without using the verifier operations abstraction. Write operations
-into the socket are rejected in the current implementation.
+pointers in a generic manner via \verb+bpf_verifier_ops+. The core of the
+verifier is not aware of specifice pointer types that are used as the context
+for the program, only how to verify the pointer access for these types.
+However, the packet event hook points targeted in this paper already contain a
+context pointer of a type that is different from \verb+bpf_sock+, so some
+additional logic was required in the verifier to understand this pointer type.
+The patch series associated with this paper introduced a new pointer type
+specific to sockets, and linked its validation directly to the aforementioned
+socket verification functions without using the verifier operations
+abstraction. Write operations into the socket are rejected in the current
+implementation. If variations in access and access type into the socket is
+desired in future based upon the hook point, then it may make sense to
+introduce a similar abstraction for the socket type.
 
@@
 The verifier is tasked with ensuring that all BPF program execution flow is
-safe to the extents where it interacts with the BPF API. In general, BPF
-programs, once verified and triggered by event hooks, will execute from the
+safe to the extents where it interacts with the BPF API. In general BPF
+programs---once verified and triggered by event hooks---will execute from the
 first instruction through the final instruction for a particular path through
-the program. Therefore, after execution of a BPF program containing resource
-acquire and release functions which are verified by the logic above, references
-to those resources should not leak.
+the program. Therefore, references to those resources should not leak if the
+program has been verified.
 
@@
 offset exceeds the length of the packet, the BPF program would be terminated
-prior to the final state. Without mitigation, this could lead to leaking of
-references at runtime, even if the instructions appear to balance acquisition
-and release at verification time. Newer BPF APIs provide better alternatives to
-these instructions for packet access, so we disallow the use of these
-instructions while holding a socket reference.
+prior to reaching the final state. Without mitigation, this could lead to
+leaking of references at runtime, even if the instructions appear to balance
+acquisition and release at verification time. Newer BPF APIs provide better
+alternatives to these instructions for packet access, so we disallow the use of
+these instructions while holding a socket reference.
 
@@
 program writers to detect the support for different protocol types at compile
-time. When a socket type is unsupported, the implementation would return
-\verb+NULL+, implying that there is no such socket. There is not a significant
-number of supported Layer~4 socket types in the Linux stack today, and the
-number is not expected to increase drastically, so it was considered simpler
-and easier to split out each Layer~4 socket lookup function into an independent
-helper call, so that the result could simply return a socket or \verb+NULL+ and
-not need to encode a representation of other error conditions.
+time. When a socket type is unsupported, the initial implementation would
+return \verb+NULL+ at runtime, implying that there is no such socket. There is
+not a significant number of supported Layer~4 socket types in the Linux stack
+today, and the number is not expected to increase drastically, so it was
+considered simpler and easier to split out each Layer~4 socket lookup function
+into an independent helper call, so that the result could simply return a
+socket or \verb+NULL+ and not need to encode a representation of other error
+conditions. This simplifies not only the API, but also the verifier
+implementation that must understand the return type of the helper functions.
 
@@
 that are not governed by RCU, the networking stack uses reference counting to
-manage socket memory instead, so the socket lookup and release functions would
-take and release references on the socket, respectively.
+manage socket memory instead, so the socket lookup and release function
+implementations must take and release references on the socket, respectively.
 
@@
 In cases where Linux is configured as a firewall between applications that
-exist in distinct Linux kernel copies running in the network, this makes sense;
-The information that is needed to perform stateful firewalling is not present
-on the Linux instance that needs this information, so the information must
-either be shared from the peers (assuming that the firewall orchestrator has
-control over the peer endpoints), or this information must be synthesised based
-upon the packets that are seen on this intermediate Linux instance. However, in
-many cases this model is used when filtering traffic where one endpoint is
-co-located with the Linux stack that performs the firewalling, which already
-contains information about the local end of the connection. In this use case,
-the connection tracking table effectively duplicates knowledge that Linux
-already creates and maintains.
+exist in distinct Linux kernel copies running in the network, this is
+intuitive. The information that is needed to perform stateful firewalling is
+not present on the Linux instance that needs this information, so to understand
+the state of the peers the information must be ether shared from the peers
+(assuming that the firewall orchestrator has control over the peer endpoints),
+or this information must be synthesised based upon the packets that are seen on
+this intermediate Linux instance. However, in many cases this model is used
+when filtering traffic where one peer is co-located with the Linux stack that
+performs the firewalling, which already contains information about the local
+end of the connection. In this use case, the connection tracking table
+effectively duplicates knowledge that Linux already creates and maintains.
 
@@
 
-\subsection{Filtering management traffic in forwarding devices}
+\subsection{Conditional forwarding of management traffic in forwarding devices}
 
@@
 the current applications without additional orchestration of BPF maps. Traffic
-which is not intended to be handled locally can then be either handled by
-additional BPF logic programmed in the XDP hook.
+which is not intended to be handled locally can then be handled by additional
+BPF logic programmed in the XDP hook.
 
@@
 socket lookup API makes the packet hooks more powerful to support use cases
-such as stateful firewalling or load-balancing based on the sockets that are
-open on a system.
+such as stateful firewalling or conditional redirect based on the sockets that
+are open on a system.
 

### 2018-11-01 paper/paper.tex
paper: Address feedback from Jarno

@@
 
-    To build such awareness into a BPF helper, the verifier needs the ability
-    to track the safety of the calls, including appropriate reference counting
-    upon the underlying socket. This paper describes extensions to the verifier
-    to perform tracking of references in a BPF program. This allows BPF
-    developers to extend the UAPI with functions that allocate and release
+    To build socket introspection into a BPF helper, the verifier needs the
+    ability to track the safety of the calls, including appropriate reference
+    counting upon the underlying socket. This paper describes extensions to the
+    verifier to perform tracking of references in a BPF program. This allows
+    BPF developers to extend the UAPI with functions that allocate and release
     resources within the execution lifetime of a BPF program, and the verifier
@@
 these implementations aim to provide the functionality of a stateful
-firewall~\cite{stateful-fw} - that is, to allow network policies to be defined
-to allow traffic from one network endpoint to talk to another network endpoint,
-and to implicitly allow responses to pass back through the firewall. Many
+firewall~\cite{stateful-fw} - that is, network policies may be defined to allow
+traffic from one network endpoint to talk to another network endpoint, and to
+implicitly allow responses to pass back through the firewall. Many
 implementations over the past decade have developed this functionality in a
@@
 the possibilities for stateful firewalling when the endpoint socket is
-co-located with the kernel providing firewalling functionality, effectively
-combining the ideas of \cite{ipfw,stateful-fw,dfw}.
+co-located with the kernel providing firewalling functionality, as is the case
+with container workloads. This effectively combines the ideas of a BPF-based
+firewall~\cite{ipfw} with the model of a stateful firewall~\cite{stateful-fw},
+operating with an understanding of socket state~\cite{dfw}.
 
@@
 
-With the verifier now aware of socket pointer types, this could be built upon
-to implicitly associate socket pointers with references to kernel resources.
+With the verifier now aware of socket pointer types, this provides a foundation
+to associate socket pointers with references to kernel resources.
 
@@
 The verifier is tasked with ensuring that all BPF program execution flow is
-safe to the extents where it interacts with the BPF API. In general BPF
+safe to the extent where it interacts with the BPF API. In general BPF
 programs---once verified and triggered by event hooks---will execute from the
@@
 not present on the Linux instance that needs this information, so to understand
-the state of the peers the information must be ether shared from the peers
+the state of the peers the information must be either shared from the peers
 (assuming that the firewall orchestrator has control over the peer endpoints),
-or this information must be synthesised based upon the packets that are seen on
-this intermediate Linux instance. However, in many cases this model is used
+or this information must be reconstructed based upon the packets that are seen
+on this intermediate Linux instance. However, in many cases this model is used
 when filtering traffic where one peer is co-located with the Linux stack that
@@
 by the actual forwarding path. Existing implementations may direct only traffic
-for a specific IP into the management CPU, or could plausibly use XDP to
-configure DoS filters so that only legitimate traffic is forwarded up the
-stack. Such implementations may not provide the degree of restriction on
-locally destined traffic that is desired, or may require additional
-co-ordination with the socket layer that requires significant orchestration.
+for a specific IP into the management CPU, or could redirect all traffic to the
+management CPU and configure DoS filters in XDP so that only legitimate traffic
+is forwarded up the stack. Such implementations may not provide the degree of
+restriction on locally destined traffic that is desired, or may require
+additional co-ordination with the socket layer that requires significant
+orchestration.
 
@@
 
-The author would like to acknowledge Alexei Starovoitov, Daniel Borkmann,
-Martin KaFai Lau, Nitin Hande and Thomas Graf for guidance, review and testing
-of the patches.
+The author would like to acknowledge Alexei Starovoitov, Daniel Borkmann, Jarno
+Rajahalme, Martin KaFai Lau, Nitin Hande and Thomas Graf for guidance, review
+and testing of the patches, and feedback on this paper.
 

### 2018-11-01 paper/paper.tex
paper: Improve use cases, conclusion

@@
 
+This work was initially motivated by an attempt to re-think stateful
+firewalling in the age of container networking, which is described in further
+detail below. During the development process, another use case was identified
+in the implementation of forwarding devices such as switches and routers, so
+more detail is provided for this use case as well. Additional use cases are
+expected to present themselves as BPF developers become more aware of the new
+capabilities.
+
 \subsection{Stateful Firewalling}
@@
 
-This paper describes the contribution of new BPF verifier functionality in the
-Linux kernel to track references to kernel resources and uses these to provide
-access to socket introspection capabilities from packet handling hooks. The
-introduction of reference tracking logic into the verifier provides
-useful base infrastructure for supporting acquire and release semantics for
-kernel resources which may prove useful for other helpers in future, and the
-socket lookup API makes the packet hooks more powerful to support use cases
-such as stateful firewalling or conditional redirect based on the sockets that
-are open on a system.
+Recent trends in the deployment of cloud applications using container images
+through orchestrators such as Kubernetes have inspired a fresh look at the
+approach used to provide networking and filtering capabilities in the network.
+In light of these, this paper looks at the ability to perform stateful
+firewalling through introspection of local kernel state. This paper describes
+the contribution of new BPF verifier functionality in the Linux kernel to track
+references to kernel resources and uses these to provide access to socket
+attributes from packet handling hooks. The introduction of reference tracking
+logic into the verifier provides useful base infrastructure for supporting
+acquire and release semantics for kernel resources which may prove useful for
+other helpers in future, and the socket lookup API makes the packet hooks more
+powerful to support use cases ranging from stateful firewalling to handling
+management traffic in network devices.
 

### 2018-11-05 paper/paper.tex
paper: Minor fixups

@@
 are within the bounds of the memory allocated for the structure, and that no
-modifications are made to the socket fields that would be cause the core socket
+modifications are made to the socket fields that would cause the core socket
 handling logic to get into a bad state. This is provided through extension of
@@
 time. When a socket type is unsupported, the initial implementation would
-return \verb+NULL+ at runtime, implying that there is no such socket. There is
-not a significant number of supported Layer~4 socket types in the Linux stack
-today, and the number is not expected to increase drastically, so it was
-considered simpler and easier to split out each Layer~4 socket lookup function
-into an independent helper call, so that the result could simply return a
-socket or \verb+NULL+ and not need to encode a representation of other error
-conditions. This simplifies not only the API, but also the verifier
-implementation that must understand the return type of the helper functions.
+return \verb+NULL+ at runtime, implying that there is no such socket. Only a
+few Layer~4 socket types exist in the Linux stack today, and the number is not
+expected to increase drastically, so it was considered simpler and easier to
+split out each Layer~4 socket lookup function into an independent helper call,
+so that the result could simply return a socket or \verb+NULL+ and not need to
+encode a representation of other error conditions. This simplifies not only the
+API, but also the verifier implementation that must understand the return type
+of the helper functions.
 
@@
 
-In cases where Linux is configured as a firewall between applications that
-exist in distinct Linux kernel copies running in the network, this is
-intuitive. The information that is needed to perform stateful firewalling is
-not present on the Linux instance that needs this information, so to understand
-the state of the peers the information must be either shared from the peers
-(assuming that the firewall orchestrator has control over the peer endpoints),
-or this information must be reconstructed based upon the packets that are seen
-on this intermediate Linux instance. However, in many cases this model is used
-when filtering traffic where one peer is co-located with the Linux stack that
+In cases where the kernel is configured as a firewall between applications that
+exist in distinct kernel instances running in the network, this is intuitive.
+The information that is needed to perform stateful firewalling is not present
+on the kernel instance that needs this information, so to understand the state
+of the peers the information must be either shared from the peers (assuming
+that the firewall orchestrator has control over the peer endpoints), or this
+information must be reconstructed based upon the packets that are seen on this
+intermediate kernel instance. However, in many cases this model is used when
+filtering traffic where one peer is co-located with the kernel stack that
 performs the firewalling, which already contains information about the local
 end of the connection. In this use case, the connection tracking table
-effectively duplicates knowledge that Linux already creates and maintains.
+effectively duplicates knowledge that the kernel already creates and maintains.
 
@@
 The author would like to acknowledge Alexei Starovoitov, Daniel Borkmann, Jarno
-Rajahalme, Martin KaFai Lau, Nitin Hande and Thomas Graf for guidance, review
-and testing of the patches, and feedback on this paper.
+Rajahalme, John Fastabend, Martin KaFai Lau, Nitin Hande and Thomas Graf for
+guidance, review and testing of the patches, and feedback on this paper.
 
